---
title: "AE05-03 Hedonic Prices and heteroscedasticity"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(alr4)
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)
library(correlation)
library(patchwork)
library(lmtest)
library(sandwich)
library(clubSandwich)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 12,
  fig.asp = 0.618,
  fig.retina = 1,
  out.width = "100%", 
  message = FALSE,
  echo = TRUE
)

my_gof <- function(fit_obj, digits = 4) {
  sum_fit <- summary(fit_obj)
  
  stars <- 
    pf(sum_fit$fstatistic[1],
       sum_fit$fstatistic[2], 
       sum_fit$fstatistic[3],
       lower.tail=FALSE) %>% 
    symnum(corr = FALSE, na = FALSE, 
           cutpoints = c(0,  .001,.01,.05,  1),
           symbols   =  c("***","**","*"," ")) %>% 
    as.character()
  
  list(
    # `R^2` = sum_fit$r.squared %>% round(digits),
    # `Adj. R^2` = sum_fit$adj.r.squared %>% round(digits),
    # `Num. obs.` = sum_fit$residuals %>% length(),
    `Num. df` = sum_fit$df[[2]],
    `F statistic` = 
      str_c(sum_fit$fstatistic[1] %>% round(digits), " ", stars)
  )
}

# Function for screening many regressors
screen_many_regs <-
  function(fit_obj_list, ..., digits = 4, single.row = TRUE) {
    
    if (class(fit_obj_list) == "lm") 
      fit_obj_list <- list(fit_obj_list)
    
    if (length(rlang::dots_list(...)) > 0)  
      fit_obj_list <- fit_obj_list %>% append(rlang::dots_list(...))
    
    # browser()
    fit_obj_list %>%
      screenreg(
        custom.note =
          map2_chr(., seq_along(.), ~ {
            str_c("Model ", .y, " ", as.character(.x$call)[[2]])
          }) %>%
          c("*** p < 0.001; ** p < 0.01; * p < 0.05", .) %>%
          str_c(collapse = "\n") ,
        digits = digits,
        single.row = single.row,
        custom.gof.rows =
          map(., ~my_gof(.x, digits)) %>%
          transpose() %>%
          map(unlist),
        reorder.gof = c(3, 4, 5, 1, 2)
      )
  }
```

# Goals:

-   Get a brief introduction to the Hedonic Prices model;
-   Practice fitting linear regression in R;
-   Test linearity assumption: visually and using a statistical tests;
-   Practice linear transformation;
-   Exerciser coefficients interpretation;

# Exercise

## 1. Load the data

```{r}
dta <- 
  alr4::MinnLand %>% 
  as_tibble()
```

## 2. Draw box-plots of `acrePrice` versus `year`

Modify variable `year` creating a factor based on `year`: use function `as.factor()`;

Play with the scale transformation of `acrePrice` to visualize data more meaningful:

-   check help for `ggplot2::scale_*_continuous(trans = "_______")`
-   check help for `ggplot2::scale_*_log10()`

```{r}
# dta %>% 
#   _________(________ = __________(________)) %>% 
#   ggplot() + 
#   aes(___ = ____, ___ = ____) + 
#   geom_________()
```

## 3. Convert monetary values from current to constant prices

Create a table with GDP deflator rates

```{r}
defl_dta <-
  tibble(
    year = 2002:2011,
    defl = c(77.47, 78.91, 81.03, 83.56, 86.09, 
             88.4, 90.12, 90.8, 91.86, 93.78)
  )
glimpse(defl_dta)
```

Join `defl_dta` to the `dta` data and create new object `dta_const`:

-   see help on joining data [here](https://rstudio-education.github.io/tidyverse-cookbook/transform-tables.html#joins);
-   convert `acrePrice` to the constant prices.
-   remove deflator variable;
-   convert variable year to a factor;

```{r}
# dta_const <- 
#   _________ %>% 
#   left_join(_________, by = "_________") %>% 
#   _________(_________ = _________ * (_________),
#             year = _________(_________)) %>% 
#   _________(- _________)
# glimpse(dta_const)
```

HW. Built a box plot out of it with log transformation.

```{r}
#
```

## 4. Produce summary statistics

Use any function of your liking. Maybe data summary skim from `modelsummary` package.

```{r}
#
```

## 5. Produce a correlation matrix

Use: `correlation::correlation()`

```{r}
# ___________ %>% 
#   ___________() %>% 
#   summary()
```

## 6. Fit regression and summarize the results

use `log(acrePrice)` as a dependent variable

```{r}
# fit1 <- lm(
#   _____ ~ _____,
#   data = _____
# )
# summary(fit1)
```

Use `performance` and `parameters` package to summarize the regression results.

```{r}
library(parameters)
library(performance)
# parameters(_____)
# performance(_____)
```

### Interpret the regression results

```{r}
# screen_many_regs(fit1)
```

## 7. Check linearity visually

```{r}
# check_model(______, check = ______)
```

## 9. Check the linearity using `car` package

```{r}
library(car)
# residualPlots(______)
```

## 11. Check multicollinearity

use `vif()` from `car`.

```{r}
# ______(______)
```

## 12. Check homoscedasticity visually

```{r}
# check_model(______, check = c("linearity", "homogeneity"))
```

## 13. Check homoscedasticity using statistical tests

```{r}
library(lmtest)
# ______(______)
```

## 14. Correct standard errors and interpret the results

```{r}
library(sandwich)
# parameters(_____, vcov = "HC3")
```

# Solutions

```{r}
#| code-fold: true
#| cache: true

## 1. Load the data
dta <- 
  alr4::MinnLand %>% 
  as_tibble() 

glimpse(dta)

## 2. Draw boxplots of `acrePrice` versus `year`
dta %>% 
  mutate(year = as.factor(year)) %>% 
  ggplot() + 
  aes(x = year,  y = acrePrice) + 
  geom_boxplot() + 
  scale_y_log10()

## 3. Convert monetary values from current to constant prices
defl_dta <-
  tibble(
    year = 2002:2011,
    defl = c(77.47, 78.91, 81.03, 83.56, 86.09, 
             88.4, 90.12, 90.8, 91.86, 93.78)
  )
glimpse(defl_dta)

dta_const <- 
  dta %>% 
  left_join(defl_dta, by = "year") %>% 
  mutate(acrePrice = acrePrice / (defl / 100),
         year = as.factor(year)) %>% 
  select(-defl)
glimpse(dta_const)

dta_const %>% 
  ggplot() + 
  aes(x = year,  y = acrePrice) + 
  geom_boxplot() + 
  scale_y_log10() + 
  labs(y = "Price per acre in constant 2010 USD")


## 4. Produce summary statistics
dta_const %>% datasummary_skim(output = "markdown")


## 5. Produce a correlation matrix
library(correlation)
dta_const %>% 
  correlation() %>% 
  summary()

## 6. Fit the basic regression and summarize the results
fit1 <- lm(
  log(acrePrice) ~ crpPct + acres + region + 
    year + tillable + productivity, 
  data = dta_const
)
summary(fit1)
parameters(fit1)
performance(fit1)


## 7. Check linearity visually
check_model(fit1, check = "linearity")

## 9. Check the linearity using `car` package

library(car)
residualPlots(fit1)

## 11. Check multicollinearity
vif(fit1)

## 12. Check homoscedasticity visually
check_model(fit1, check = c("linearity", "homogeneity"))

## 13. Check homoscedasticity using statistical tests
library(lmtest)
bptest(fit1)

## 14. Correct standard errors
library(lmtest)
library(sandwich)
parameters(fit1, vcov = "HC3")
```
