---
title: "AE06-02 Interaction terms with continuous variables"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(alr4)
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)
library(correlation)
library(patchwork)
library(lmtest)
library(sandwich)
library(clubSandwich)
library(forcats)
library(modelbased)
library(emmeans)
library(ggeffects)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 16,
  fig.asp = 0.618,
  fig.retina = 1,
  out.width = "100%", 
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
```

# Goals:

-   Learn what is the interaction term in regression and how to use it;


# Exercise 1. Interaction with a continious variables

## 1.1 Load data, compute weekdays

```{r}
set.seed(1223)
icecream <- 
  read_csv("data/ice_cream_sales_rnd.csv") %>% 
  mutate(
    weekday = case_when(
      weekday == 1 ~ "Sun",
      weekday == 2 ~ "Mon",
      weekday == 3 ~ "Tue",
      weekday == 4 ~ "Wed",
      weekday == 5 ~ "Thu",
      weekday == 6 ~ "Fri",
      weekday == 7 ~ "Sat") %>% 
      as_factor()
  ) %>% 
  sample_n(1500)
```

## 1.2 Describe the data

```{r}
library(GGally)
ggpairs(icecream)
icecream %>% datasummary_skim()
```

## 1.3 Fit the model with an interaction between temperature and 

Built a regression model with an interaction terms between prices, weekdays and temperatures.

```{r}
fit1 <- lm(log(sales) ~ temp * price * weekday, icecream)
parameters(fit1)
performance(fit1)
```

## 1.4 Interprete the interaction term between prices and temperature

`estimate_slopes` and `ggpredict`
```{r}
# estimate_slopes(______, trend = ______, at = ______ ) %>% plot()
# ggpredict(______, terms = c(______, "temp [10, 25, 35]") ) %>% ______()
```


## 1.5 Interprete the interaction term between prices and temperature

```{r}
#
#
#
```
