---
title: "AE06-01 linear transformation and interaction term: the toolbox"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(alr4)
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)
library(correlation)
library(patchwork)
library(lmtest)
library(sandwich)
library(clubSandwich)
library(forcats)
library(modelbased)
library(emmeans)
library(ggeffects)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 16,
  fig.asp = 0.618,
  fig.retina = 1,
  out.width = "100%", 
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
```

# Goals:

-   Learn what is the interaction term in regression and how to use it;


# Exercise 1. Interaction with a dummy variable

## 1.1 Load data, compute, select, filter

-   total amount of parents education `paredu` as a sum of years of their education

```{r}
dta <- 
  wooldridge::wage1 %>% 
  as_tibble() %>% 
  mutate(sex = ifelse(female == 0, "male", "female") %>% 
           as_factor() %>% 
           fct_rev()) %>%
  select(wage, educ, exper, sex)
```

Variables are:

-   `wage`: average hourly earnings
-   `educ`: years of education
-   `exper`: years potential experience
-   `sex`: sex of the respondents

## 1.2 Built pairs plot and descriptive statistics

```{r}
library(GGally)
ggpairs(dta, aes(group = sex, colour = sex, alpha = 0.3))
```

```{r}
library(modelsummary)
dta %>% datasummary_skim()
```

## 1.3 Regression without interaction terms

```{r}
fit_1 <- lm(wage ~ educ + exper + sex, data = dta)
parameters(fit_1)
performance(fit_1)
```

### Diagnostics

```{r}
check_model(fit_1, check = c("linearity", "homogeneity"))
```

## 1.4 Attempting to resolve linearity

- Introduce squared education and experience terms.

```{r}
# fit_2 <- lm(log(wage) ~ educ + ______ + exper + ______ + sex, 
#             data = dta)
# parameters(fit_2)
# performance(fit_2)
# check_model(fit_2, check = c("linearity", "homogeneity"))
```

## 1.5 Computing fitted values

Calculate fitted values based on the regression`fit_2` for: 

-   a women and men;
-   with  10 and 11 years of experience;
-   10, 11, 16 and 17 years of education;

Built a `tibble()` with the covariates for prediction:

```{r}
# pred_dta_1 <-
#   tibble(
#     sex = ______,
#     ______ = c(10, 11, ______, 10, 11, 16, 17),
#     exper = ______
#   ) %>%
#   bind_rows(mutate(., sex = "male"))
```

Predict fitted values and write them to the variable fitted

```{r}
# ______(______, pred_dta_1)
# 
# pred_dta_2 <- 
#   pred_dta_1 %>% 
#   mutate(fitted_values = ______(______, pred_dta_1))
```

Compute predicted salary and save it into the variable `pred_wage`;

```{r}
# pred_dta_3 <- 
#   pred_dta_2 %>% 
#   mutate(pred_wage = ______(fitted_values))
# pred_dta_3
```

## 1.6 Computing slope for squared term

### The slope depends on the value of regressor

$\frac{\partial \text{ wage}}{\partial \text{ educ}} = -0.0306529 + 2 \times 0.0048716 \times \text{educ}$

What is the effect of an additional year of education on wage, given that current level of education is 10 or 16 years?

-   Estimate slopes manually and using `modelbased`;
-   Interpret the results;
-   Plot the slope;
-   What is the meaning of the shaded area?

```{r}
library(modelbased)
library(emmeans)

## Estimate slope of educ for 10 and 11 years of education, manually
# parameters(fit_2)
# ______ + ______ * ______ * ______
# ______ + ______ * ______ * ______

# # Estimating slopes manually for our data
# pred_dta_3 %>% 
#   select(- pred_wage  ) %>%
#   pivot_wider(names_from = educ, values_from = fitted_values) %>% 
#   mutate(beta_educ_10_11 = `11`-`10`,
#          beta_educ_16_17 = `17`-`16`)

# # Using function estimate_slopes
# estimate_slopes(fit_2, trend = "educ")
# 
# # add `at = "educ"`
# estimate_slopes(fit_2, trend = ______, at = "educ")
# 
# # Add `at = list(educ = c(10, 11))`
# estimate_slopes(fit_2, trend = ______, at = list(educ = c(10, 11)))
# 
# # adding plot
# estimate_slopes(fit_2, trend = ______, at = ______) %>% plot()
# 
# # adding plot at sex and education
# estimate_slopes(fit_2, trend = ______, at = c("educ", "sex")) %>% ______()

```

### Marginal effects to report

## 1.7 Predicted values given linear transformation

```{r}
library(ggeffects)

# pred_dta_3
# 
# # using ggpredict
# ggpredict(fit_2, terms = "educ")
# 
# # back.transform = F
# ggpredict(fit_2, ______ = "educ", back.transform = F)
# 
# # `terms = c("educ", "sex")`
# ggpredict(fit_2, terms = c("educ", "sex"), ______)
# 
# # Plotting
# ggpredict(fit_2, ______ = "educ", back.transform = F) %>% plot()
# ggpredict(fit_2, ______) %>% plot()
# ggpredict(fit_2, terms = c(______)) %>% plot()
```


## 1.8 Effect of a dummy variable

In the regression coefficients, sex variable has a coefficient of -0.32

```{r}
# ggpredict(fit_2, ______ = "sex") %>% plot()
# ggpredict(fit_2, ______ = "sex", back.transform = F)
# parameters(fit_2)
```

# Exercise 2. Interaction term


## 2.1 Adding an interaction term between sex and education

```{r}
# fit_3 <- lm(log(wage) ~ 
#               educ + I(educ ^ 2) + exper + I(exper ^ 2) + sex + 
#               ______, 
#             data = dta)
# parameters(fit_3)
# performance(fit_3)
# check_model(fit_3, check = c("linearity", "homogeneity"))
```

## 2.2 Interpreting the interaction term

Interpreting the changing slopes

```{r}
# estimate_slopes(______, ______ = "educ", at = "educ") %>% plot()
# estimate_slopes(______, ______ = "educ", ______ = c("sex", "educ")) %>% plot()
# estimate_slopes(______, ______ = "educ", ______ = c("sex")) %>% plot()
```

Interpreting the fitted values

```{r}
# ggpredict(______, ______ = "educ") %>% plot()
# ggpredict(______, ______ = c("sex", "educ")) %>% plot()
# ggpredict(______, ______ = c("educ", "sex")) %>% plot()
# ggpredict(______, ______ = c("educ", "sex")) %>% plot(add.data = TRUE)
```



# Solutions 

```{r}
#| code-fold: true
dta <- 
  wooldridge::wage1 %>% 
  as_tibble() %>% 
  mutate(sex = ifelse(female == 0, "male", "female") %>% 
           as_factor() %>% 
           fct_rev(),
         lwage = log(wage)) %>%
  select(lwage, wage, educ, exper, sex)

# 1.2 Built pairs plot and descriptive statistics
library(GGally)
ggpairs(dta, aes(group = sex, colour = sex, alpha = 0.3))

library(modelsummary)
dta %>% datasummary_skim()

# 1.3 Regression without interaction terms
fit_1 <- lm(wage ~ educ + exper + sex, data = dta)
parameters(fit_1)
performance(fit_1)
check_model(fit_1, check = c("linearity", "homogeneity"))

# 1.4 Attempting to resolve linearity
fit_2 <- lm(log(wage) ~ educ + I(educ ^ 2) + exper + I(exper ^ 2) + sex, 
            data = dta)
parameters(fit_2)
performance(fit_2)
check_model(fit_2, check = c("linearity", "homogeneity"))

# 1.5 Computing fitted values
pred_dta_1 <-
  tibble(
    sex = "female",
    educ = c(10, 11, 16, 17, 10, 11, 16, 17),
    exper = c(10, 10, 10, 10, 11, 11, 11, 11)
  ) %>%
  bind_rows(mutate(., sex = "male"))

# Predict fitted values and write them to the variable fitted
pred_dta_2 <- 
  pred_dta_1 %>% 
  mutate(fitted_values = predict(fit_2, pred_dta_1))


# Compute predicted salary and save it into the variable `pred_wage`;
pred_dta_3 <- 
  pred_dta_2 %>% 
  mutate(pred_wage = exp(fitted_values))
pred_dta_3

# 1.6 Computing slope for squared term
library(modelbased)
library(emmeans)

# Estimate slope of educ for 10 and 11 years of education, manually
-0.0306529 + 0.0048716 * 2 * 10
-0.0306529 + 0.0048716 * 2 * 11

# Estimating slopes manually for our data
pred_dta_3 %>% 
  # filter(educ %in% c(10, 11)) %>% 
  select(- pred_wage  ) %>%
  pivot_wider(names_from = educ, values_from = fitted_values) %>% 
  mutate(beta_educ_10_11 = `11`-`10`,
         beta_educ_16_17 = `17`-`16`)

# Using function estimate_slopes
estimate_slopes(fit_2, trend = "educ")

# add `at = "educ"`
estimate_slopes(fit_2, trend = "educ", at = "educ")

# Add `at = list(educ = c(10, 11))`
estimate_slopes(fit_2, trend = "educ", at = list(educ = c(10, 11)))

# adding plot
estimate_slopes(fit_2, trend = "educ", at = "educ") %>% plot()

# adding plot at sex and education
estimate_slopes(fit_2, trend = "educ", at = c("educ", "sex")) %>% plot()

# 1.7 Predicted values given linear transformation
library(ggeffects)

pred_dta_3

# using ggpredict
ggpredict(fit_2, terms = "educ")

# back.transform = F
ggpredict(fit_2, terms = "educ", back.transform = F)

# `terms = c("educ", "sex")`
ggpredict(fit_2, terms = c("educ", "sex"), back.transform = F)

# Plotting
ggpredict(fit_2, terms = "educ", back.transform = F) %>% plot()
ggpredict(fit_2, terms = "educ") %>% plot()
ggpredict(fit_2, terms = c("educ", "sex")) %>% plot()

# 1.8 Effect of a dummy variable
ggpredict(fit_2, terms = "sex") %>% plot()
ggpredict(fit_2, terms = "sex", back.transform = F)
parameters(fit_2)

# Exercise 2. Interaction term
## 2.1 Adding an interaction term between sex and education
fit_3 <- lm(log(wage) ~ 
              educ + I(educ ^ 2) + exper + I(exper ^ 2) + sex + 
              educ:sex, 
            data = dta)
parameters(fit_3)
performance(fit_3)
check_model(fit_3, check = c("linearity", "homogeneity"))

## 2.2 Interpreting the interaction term
estimate_slopes(fit_3, trend = "educ", at = "educ") %>% plot()
estimate_slopes(fit_3, trend = "educ", at = c("sex", "educ")) %>% plot()
estimate_slopes(fit_3, trend = "educ", at = c("sex")) %>% plot()

ggpredict(fit_3, terms = "educ") %>% plot()
ggpredict(fit_3, terms = c("sex", "educ")) %>% plot()
ggpredict(fit_3, terms = c("educ", "sex")) %>% plot()
ggpredict(fit_3, terms = c("educ", "sex")) %>% plot(add.data = TRUE)
```
