---
title: "Introduction. Organisation. Setup."
subtitle: "MP223 - Applied Econometrics Methods for the Social Sciences"
author: "Eduard Bukin"
studIP: "[StudIP profile](https://studip.uni-giessen.de/dispatch.php/profile?cid=b1775c5097d2a85c2986b7667d22c3dc&username=gh2910)"
email: "[eduardbukin@agrar.uni-giessen.de](mailto:eduardbukin@agrar.uni-giessen.de)"
phone: "+49 641 99-37055"
office: "Zeughaus (Senckenbergstr. 3). Room: 132"
hours: "Part time (better to make an appointment)"
footer:  "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "../../img/jlu-logo.png"
editor: visual
format: 
  revealjs: 
    transition: fade
    slide-number: true
    smaller: true
    scrollable: true
execute:
  freeze: auto
---

```{r}
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  out.width = "80%"
)
```

## `r rmarkdown::metadata$subtitle`

**Author**: `r rmarkdown::metadata$author` (`r rmarkdown::metadata$studIP`)

**Email**: `r rmarkdown::metadata$email`,

**Telephone**: `r rmarkdown::metadata$phone`

**Office**: `r rmarkdown::metadata$office`.

**Office hours**: `r rmarkdown::metadata$hours`

## Introduction

Welcome to the `r rmarkdown::metadata$subtitle`!

## Lecturers: **Christoph Funk** {.smaller}

[Christoph.Funk\@wirtschaft.uni-giessen.de](mailto:Christoph.Funk@wirtschaft.uni-giessen.de). [Website](https://www.uni-giessen.de/fbz/zentren/zeu/Team/zeumembersall/Christoph_Funk/dpfunk).

Post Doc.

> Center for international Development and Environmental Research (ZEU) Justus Liebig Universität

2020 - PhD in economics from Justus Liebig University Giessen

Research interests: - SDG monitoring - Climate change vulnerability - Adaptation strategies - Energy economics - Econometric modelling

## Lecturers: **Vladimir Otrachshenko** {.smaller}

[Vladimir.Otrachshenko\@zeu.uni-giessen.de](mailto:Vladimir.Otrachshenko@zeu.uni-giessen.de). [Website](https://www.uni-giessen.de/fbz/zentren/zeu/Team/zeumembersall/Vladimir_Otrachshenko/dpotrachshenko).

Senior Researcher.

> Center for international Development and Environmental Research (ZEU) Justus Liebig Universität

2013 - PhD in Economics from Nova School of Business and Economics, Lisbon, Portugal

Research interests:

-   Environmental and Resource Economics
-   Climate Change
-   Health and Population Economics

## Lecturers: **`r rmarkdown::metadata$author`** {.smaller}

`r rmarkdown::metadata$email`, (`r rmarkdown::metadata$studIP`)

Data science enthusiast, econometrics practitioner. PhD Student.

> Institute of Agricultural Policy and Market Research

2015 -- MSc in Rural Development:

-   Ghent University, Belgium

Research interests:

-   Restructuring and productivity change in agriculture
-   Land and labor factor markets in agriculture

## Your turn!

**Please introduce yourself**

. . .

**What is your name?**

**Where do you come from?**

. . .

**What do you study?**

**What is your background?**

. . .

**What are your expectations?**

## Course structure and overview (1/4)

`r rmarkdown::metadata$subtitle` is taught **in presence**.

::: incremental
-   Every Wednesday 14:00 - 18:00, Room: Senckenbergstr. 03, 216 (Ze-PC2)

-   Wear a mask all the time.

-   Get a new mask from a lecturer for every lecture.

-   Optional, make a COVID-19 speed tests if you do no feel well (ask the test from a lecturer).

-   Please, do not show up if you are sick.
:::

## Course structure and overview (2/4) {.smaller}

. . .

Online resources:

-   [Ilias](https://ilias.uni-giessen.de/goto.php?target=crs_301421&client_id=JLUG) is used for materials dissemination;

-   [StudIP](https://studip.uni-giessen.de/dispatch.php/course/overview?cid=b1775c5097d2a85c2986b7667d22c3dc) is only used for announcements.

. . .

Rough course structure is in the **Course Plan on Ilias.**

. . .

Almost every week will have a **checklist**

-   summary of materials from the class;

-   key materials to cover on your own;

## Course structure and overview (3/4) {.smaller}

**Lectures**

-   Slides on Ilias. No pre-recording;
-   "Takeaways" slides in th end of a lecture (if available);

. . .

**Application exercises** in class

-   Sometimes have pre requisites (watch a video, read a paper)
-   Sometimes require preparation in advance;

. . .

-   **Seminar discussions** - definitely require preparation (reading papers) in advance;

. . .

**Application exercises at home**

-   Some are mandatory, some are optional but highly recommended.
-   With /or without video guidance.

## Course structure and overview (4/4)

Examination:

-   60% written exam

. . .

-   40% practical homework on R analysis.

    -   Will be disseminated on the week 4-6.
    -   Deadline: end of semester.

## Any questions?

## Plan for today

Next 45 minutes: Application Exercise 01 - Soft introduction to R

-   in class we will do "ae01a-vaccination.Rmd" everything else is at home;

. . .

Two Lectures (90 min):

-   Ceteris Paribus
-   Selection Bias

. . .

Application Exercise at home.

-   See the check list.

## Application Exercise 01 - Soft introduction to R

. . .

**Turn on your PC**

::: callout-important
Login: `ZH-user-pcl`

Password: `V5-senc!3ken`
:::

. . .

Write them down and remember.

Same password an login will be used on all PCs in Zeughaus.

## Log into studIP and Ilias

. . .

Log into your `studIP`;

. . .

Follow to "Ilias \>\> Kurs (ID 301421) in JLUG";

```{r}
#| echo: false
#| out.width: "100%"
#| fig.asp: 0.2

library(magrittr)
here::here("slides", "00-introduction", "img", "ptudip-ilias.PNG") %>% 
  magick::image_read() %>%
  magick::image_ggplot(interpolate = TRUE)
```

. . .

Log in to Ilias;

. . .

Download `ae01-soft-intro-to-R.zip` to downloads;

## Setup working folders

. . .

Navigate to your user folder: `C > Users > Name of your user account`;

. . .

-   Create there a **course folder** names `{your initial}-mk223-2022`.

    -   Use it for your course for all in-class work;
    -   on my pc the course folder is called `eb-mk223-2022`;
    -   the full path is `C:\Users\ZH-user-pcl\eb-mk223-2022`;

. . .

-   Paste `ae01-soft-intro-to-R.zip` from downloads to the **course folder**;

. . .

-   Unzip `ae01-soft-intro-to-R.zip` into `ae01-soft-intro-to-R`;

## Launch the R Studio from the project "ae01-soft-intro-to-R"

. . .

-   Navigate to `ae01-soft-intro-to-R` in your **course folder**

. . .

-   Open `ae01-soft-intro-to-R.Rproj` that has R studio icon and `.Rproj` extension:

::: columns
::: {.column width="50%"}
```{r}
#| echo: false
#| out.width: "45%"
#| fig.asp: 1
#| fig.retina: 
here::here("slides", "00-introduction", "img", "RStud-icon.PNG") %>% 
  magick::image_read() %>%
  magick::image_ggplot(interpolate = TRUE)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| out.width: "100%"
#| fig.asp: .05
#| fig.dpi: 300

here::here("slides", "00-introduction", "img", "RStud-icon-2.PNG")%>% 
  magick::image_read() %>%
  magick::image_ggplot(interpolate = TRUE)
```
:::
:::
