---
title: "Hedonic model"
subtitle: "MP223 - Applied Econometrics Methods for the Social Sciences"
author: "Eduard Bukin"
footer:  "[https://ebukin.github.io/mp223-2022-aem-R-public/](https://ebukin.github.io/mp223-2022-aem-R-public/)"
logo: "../../img/jlu-logo.png"
editor: visual
format: 
  revealjs: 
    transition: fade
    slide-number: true
    smaller: false
    scrollable: true
    
execute:
  freeze: auto
bibliography: ../../references.bib
---

# R setup {.smaller}

```{r}
#| echo: true

library(tidyverse)       # for data wrangling
library(alr4)            # for the data sets #
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)

ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.width = 10,
  fig.asp = 0.5,
  fig.retina = 3,
  dpi = 300,
  out.width = "100%", 
  message = FALSE,
  echo = TRUE, 
  cache = TRUE
)


my_gof <- function(fit_obj, digits = 4) {
  sum_fit <- summary(fit_obj)
  
  stars <- 
    pf(sum_fit$fstatistic[1],
       sum_fit$fstatistic[2], 
       sum_fit$fstatistic[3],
       lower.tail=FALSE) %>% 
    symnum(corr = FALSE, na = FALSE, 
           cutpoints = c(0,  .001,.01,.05,  1),
           symbols   =  c("***","**","*"," ")) %>% 
    as.character()
  
  list(
    # `R^2` = sum_fit$r.squared %>% round(digits),
    # `Adj. R^2` = sum_fit$adj.r.squared %>% round(digits),
    # `Num. obs.` = sum_fit$residuals %>% length(),
    `Num. df` = sum_fit$df[[2]],
    `F statistic` = 
      str_c(sum_fit$fstatistic[1] %>% round(digits), " ", stars)
  )
}

screen_many_regs <-
  function(fit_obj_list, ..., digits = 4, single.row = TRUE) {
    
    if (class(fit_obj_list) == "lm") 
      fit_obj_list <- list(fit_obj_list)
    
    if (length(rlang::dots_list(...)) > 0)  
      fit_obj_list <- fit_obj_list %>% append(rlang::dots_list(...))
    
    # browser()
    fit_obj_list %>%
      screenreg(
        custom.note =
          map2_chr(., seq_along(.), ~ {
            str_c("Model ", .y, " ", as.character(.x$call)[[2]])
          }) %>%
          c("*** p < 0.001; ** p < 0.01; * p < 0.05", .) %>%
          str_c(collapse = "\n") ,
        digits = digits,
        single.row = single.row,
        custom.gof.rows =
          map(., ~my_gof(.x, digits)) %>%
          transpose() %>%
          map(unlist),
        reorder.gof = c(3, 4, 5, 1, 2)
      )
  }


```

# Problem Setting {.smaller}

We would like to assess the effect of the "Conservation Reserve Program" (CPR) on the agricultural land prices in Minnesota in 2002-2011.

. . .

Conservation Reserve Program

-   is a subsidy

-   obligates farms NOT TO GROW ANY CROPS on the enrolled land

-   pays monetary compensation in exchange;

. . .

As this is a risk less income, it may increase the land price.

. . .

However, if the land is productive, such set-aside measure may reduce farmland price.

## Hedonic Model overview

Hedonic prices is an econometric approach of quantifying monetary values of differentiated characteristics of goods and services, which are subjects of economic exchange.

. . .

For example, agricultural land.

-   It has such characteristics as: location, slope, environmental limitation, farmers' accessibility, climate, expected rainfall, soil salinity, nutrient content, irrigation availability and other.

## Hedonic Theory (1/2) {.smaller}

According to [@palmquist1989], hedonic model of agricultural land price is based on the equilibrium between **Offer** and **Bid** functions. Author uses partial equilibrium approach to prove this.

. . .

$$\phi(\hat{z}, \tilde{z}, \pi^{S^{'}}, r, \beta) = R = \pi^{S^{'}} + C(\hat{z}, \tilde{z}, r, \beta)$$

. . .

$R(\cdot)$ - realized land price (rental of sales);

$\hat{z}$ - land characteristics exogenous to land owner;

$\tilde{z}$ - land characteristics in control of land owner;

$r$ - inputs prices;

$\beta$ - technologies and opportunities such as credit availability;

$\pi^{S^{'}}$ - expected profit of agricultural producers from land;

## Hedonic Theory (2/2)

Essentially, **causal function** of the land rental price could be written as:

$$R = R(\hat{z}, \tilde{z}, \pi^{S^{'}}, r, \beta)$$

. . .

::: callout-important
This is a causal relationship because [@palmquist1989] provides solid theoretical justification for it.
:::

## What are the differentiated land characteristics? {.smaller}

### Affected by land owner:

-   a\.
-   b\.
-   c\.

### Not affected by land owner:

-   a\.
-   b\.
-   c\.

# Data {.smaller}

Data set `MinnLand` from [@taff2007compensated].

::: {.incremental}

-   `acrePrice` - sale price in dollars per acre;
-   `acres` - size of the farm in acres;
-   `region` - region in the state Minnesota;
-   `year` - year of the land sales translation;
-   `crpPct` - the percentage of all farm acres enrolled in CRP;
-   `tillable` - percentage of farm acreage that is rated arable by the assessor;
-   `productivity` - average agronomic productivity scaled 1 to 100, with larger numbers for more productive land;

:::

## Loading data

```{r}
#| code-fold: true
dta <- 
  alr4::MinnLand %>% 
  as_tibble()  %>%
  # filter(year == 2007) %>% 
  select(acrePrice, acres, region, year, 
         tillable, crpPct, productivity)
glimpse(dta)
```

## Summary statistics (1/2)

```{r}
#| code-fold: true
dta %>%
  mutate(id = row_number()) %>% 
  pivot_longer(
    cols = c(acrePrice, acres, year, tillable, crpPct, productivity),
    names_to = "var",
    values_to = "val"
  ) %>% 
  group_by(var) %>% 
  summarise(across(
    c(val),
    list(
      mean = ~ mean(.x, na.rm = TRUE),
      sd = ~ sd(.x, na.rm = TRUE),
      meadian = ~ median(.x, na.rm = TRUE),
      n_miss = ~ sum(is.na(.x), na.rm = TRUE),
      min = ~ min(.x, na.rm = TRUE),
      max = ~ max(.x, na.rm = TRUE)
    )
  ), 
  n = n())
```

## Summary statistics (2/2) {.smaller}

```{r}
#| code-fold: true
dta %>% datasummary_skim(output = "data.frame")
```

## Visual inspection: scatter plots {.smaller}

```{r}
#| code-fold: true
#| out-width: 100%
dta %>%
  select(acrePrice, acres, tillable, 
         crpPct, productivity) %>%
  ggpairs(aes(alpha = 0.5)) +
  theme_bw()
```

# Hedonic Regression

## Regression. Summary {.smaller}

```{r}
#| code-fold: true
fit1 <- lm(
  acrePrice ~ crpPct + acres + region + 
    year + tillable + productivity, 
  data = dta
)
summary(fit1)
```

## Regression. Interpretation of the coefficients {.smaller}

```{r}
parameters(fit1)
```

## Checking Linearity 1/3

```{r}
check_model(fit1, check = "linearity", panel = FALSE)
```

## Checking Linearity 2/3

```{r}
residualPlots(fit1, plot = FALSE)
```


## Checking Linearity 3/3

```{r}
#| code-fold: true
residualPlots(fit1, test = FALSE, layout = c(2, 2), ask = FALSE)
```

# Takeaways

## Takeaway {.smaller}

-   Multiple linear regression
-   Hedonic model
-   Slope and Intercept (interpretation)
-   Residuals vs Fitted

# Homework

Create an R Script out of the R code in the presentation.

# References

## References
