---
title: "Omitted Variable Bias"
subtitle: "MP223 - Applied Econometrics Methods for the Social Sciences"
author: "Eduard Bukin"
footer:  "[https://ebukin.github.io/mp223-2022-aem-R-public/](https://ebukin.github.io/mp223-2022-aem-R-public/)"
logo: "../../img/jlu-logo.png"
editor: visual
format: 
  revealjs: 
    transition: fade
    slide-number: true
    smaller: false
    scrollable: true
    
execute:
  freeze: auto
bibliography: ../../references.bib
---

# R setup {.smaller}

```{r}
#| echo: true

library(tidyverse)       # for data wrangling
library(alr4)            # for the data sets #
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)

ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.width = 10,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "100%", 
  message = FALSE,
  echo = TRUE, 
  cache = TRUE
)


my_gof <- function(fit_obj, digits = 4) {
  sum_fit <- summary(fit_obj)
  
  stars <- 
    pf(sum_fit$fstatistic[1],
       sum_fit$fstatistic[2], 
       sum_fit$fstatistic[3],
       lower.tail=FALSE) %>% 
    symnum(corr = FALSE, na = FALSE, 
           cutpoints = c(0,  .001,.01,.05,  1),
           symbols   =  c("***","**","*"," ")) %>% 
    as.character()
  
  list(
    # `R^2` = sum_fit$r.squared %>% round(digits),
    # `Adj. R^2` = sum_fit$adj.r.squared %>% round(digits),
    # `Num. obs.` = sum_fit$residuals %>% length(),
    `Num. df` = sum_fit$df[[2]],
    `F statistic` = 
      str_c(sum_fit$fstatistic[1] %>% round(digits), " ", stars)
  )
}

screen_many_regs <-
  function(fit_obj_list, ..., digits = 4, single.row = TRUE) {
    
    if (class(fit_obj_list) == "lm") 
      fit_obj_list <- list(fit_obj_list)
    
    if (length(rlang::dots_list(...)) > 0)  
      fit_obj_list <- fit_obj_list %>% append(rlang::dots_list(...))
    
    # browser()
    fit_obj_list %>%
      screenreg(
        custom.note =
          map2_chr(., seq_along(.), ~ {
            str_c("Model ", .y, ": ", as.character(.x$call)[[2]])
          }) %>%
          c("*** p < 0.001; ** p < 0.01; * p < 0.05", .) %>%
          str_c(collapse = "\n") ,
        digits = digits,
        single.row = single.row,
        custom.gof.rows =
          map(., ~my_gof(.x, digits)) %>%
          transpose() %>%
          map(unlist),
        reorder.gof = c(3, 4, 5, 1, 2)
      )
  }
```

# Omitted Variable Bias

In multiple regression, Ceteris Paribus is achieved by introducing control variables.

. . .

::: callout-warning
Having insufficient or not right controls leaves us with the Selection Bias, which is called **OVB - Omitted Variable Bias** in the regression analysis.
:::

## Omitted Variable {.smaller}

#### Long Model

$$Y_i = \alpha ^ l + \beta ^ l P_i + \gamma A_i + e^l_i$$

where:

::: incremental
-   $Y_i$ is the outcome variable;
-   $P_i$ is the key variable of treatment effect;
-   $\alpha ^ l$ , $\beta ^ l$ are true regression coefficients;
-   $\gamma$ is the effect of omitted variable in long;
-   $A_i$ is the **omitted variable**;
:::

## Omitted Variable {.smaller}

#### Long Model

$$Y_i = \alpha ^ l + \beta ^ l P_i + \gamma A_i + e^l_i$$

where:

-   $Y_i$ is the outcome variable;
-   $P_i$ is the key variable of treatment effect;
-   $\alpha ^ l$ , $\beta ^ l$ are true regression coefficients;
-   $\gamma$ is the effect of omitted variable in long;
-   $A_i$ is the **omitted variable**;

#### Short model

$$Y_i = \alpha ^ s + \beta^s P_i + e^s_i$$

Is the model that omit one or more variables compare to the long one.

## OVB logic

------------------------------------------------------------------------

### Omitted Variable Bias (1/3)

Omitted variable $A_i$ causes **bias** if:

::: incremental
1.  ::: {.callout-important appearance="simple"}
    $P_i$ and $A_i$ relates to each other:

    -   $E[A_i|P_i] \neq 0$ ; or
    -   in a regression $A_i = \pi_0 + \pi_1 P_i + u$ , $\pi_1 \neq 0$ ;
    :::

2.  ::: {.callout-important appearance="simple"}
    $A_i$ and $Y_i$ relates to each other:

    -   $E[Y_i| A_i] \neq 0$ in the long regression or $\gamma \neq 0$);
    :::
:::

$A_i = \pi_0 + \pi_1 P_i + u$ is an auxiliary regression because it helps us understand the true causal relationship.

------------------------------------------------------------------------

### Omitted Variable Bias (2/3)

$\textit{Biased effect of treatment in short} = \\ \textit{True effect of treatment in long} + \\ ([\textit{Relationship between OV and treatment}] \times \\ [\textit{Effect of OV in long}])$

------------------------------------------------------------------------

### Omitted Variable Bias (3/3)

$\textit{OVB} = \\ \beta^s - \beta^l = \pi_1 \times \gamma = \\ \textit{Regression of omitted on included } \times \\ \textit{Effect of Omitted Variable in long}$

-   $\beta^s$ is the biased effect of treatment in short;
-   $\beta^l$ is the true effect of treatment in long;
-   $\pi_1^l$ is the coef. of treatment in the auxiliary of omitted on included (indep. in short);
-   $\gamma$ is the coef. of Omitted variable in long regression;

## Why OVB formula is important? {.smaller}

::: incremental
-   **Omitted Variable** - means that we cannot have it in the regression!

-   We can't use data to check the consequences of omitting variables that we don't observe.

-   Having knowledge of mathematics behind OVB, we can **make an educated guess about consequences of variable omission: the BIAS** [@Angrist2014]

-   For example, we can write our:

    -   auxiliary regression: $A_i = \pi_0 + \pi_1 P_i + u$ ; and
    -   potential regression: $Y_i = \alpha ^ l + \beta^l P_i + \gamma A_i + e^l_i$

-   because $\text{OVB} = \pi_1 \times \gamma$

    -   we can speculate about signs of $\pi_1$ and $\gamma$
    -   we can justify how omitted variable biases our regression: upwards (increasing the effect of interest) or downwards (decreeing it).
:::

## How to resolve the OVB?

-   No solution!
    -   Find proxies
    -   Develop an elaborate research design
-   Acknowledge presence of the OVB
-   Discuss the extent of bias

# Example 1: Empirical proof of the OVB

## MLR: fertility rates and development

We explore `UN11` data from [@weisberg2005applied]

-   It has 199 observations and 6 variables

-   Variables are:

    -   `fertility` - number of children per woman;
    -   `lifeExpF` - Female life expectancy, years;
    -   `ppgdp` - Per capita gross domestic product in US dollars;
    -   `pctUrban` - Percent of Urban population;
    -   `group` - variable with 3 values "oecd", "africa" and "others" to indicate if the country belonged respectively to the Organization for Economic Co-operation and Development, African continent, and other countries.
    -   `africa`, `other` and `oecd` - dummy variables taking values of 1 if a country is in respectively Africa, other countries or OECD.

## Research question

How does the life expectancy affect fertility?

------------------------------------------------------------------------

### Empirical model

. . .

$\hat{fertility} = f(\textit{lifeExpF}, \textit{pctUrban}, \textit{group})$

. . .

This is, however, incomplete model, as we intentionally omitted variable $\textit{ppgdp}$, which might be very important.

. . .

The long model should actually be $\hat{fertility} = f(\textit{lifeExpF}, \textit{pctUrban}, \textit{group}, \textit{ppgdp})$

## Data

```{r}
#| code-fold: true
library(alr4)           #install.packages("alr4")
library(tidyverse)      #install.packages("tidyverse")
un_dta <- 
  alr4::UN11 %>%
  as_tibble() %>%
  select(-region) %>%
  mutate(
    Africa = as.integer(group  == "africa"),
    Other = as.integer(group  == "other"),
    OECD = as.integer(group  == "oecd")
  )
glimpse(un_dta)
```

------------------------------------------------------------------------

### Descriptive statistics {.smaller}

```{r}
#| code-fold: true
un_dta %>% datasummary_skim(output = "markdown")
```

------------------------------------------------------------------------

### Data visualization

```{r}
#| code-fold: true
#| fig-height: 5
#| fig-asp: .5
un_dta %>% 
  select(fertility, lifeExpF, pctUrban, ppgdp, group) %>%
  ggpairs(aes(alpha = 0.6))
```

------------------------------------------------------------------------

### Data visualization by group

```{r}
#| code-fold: true
#| fig-height: 5
#| fig-asp: .5
un_dta %>% 
  select(fertility, lifeExpF, pctUrban, ppgdp, group) %>%
  ggpairs(aes(colour = group, alpha = 0.6))
```

## Regression

### Short regression - important control variable is omitted

```{r}
#| code-fold: true
fit_s <- lm(fertility ~ lifeExpF + pctUrban + group , un_dta)
parameters(fit_s)
```

------------------------------------------------------------------------

### Educated guess: effect of the omitted variable (1/3)

::: columns
::: {.column width="50%"}
```{r}
#| code-fold: true
screen_many_regs(fit_s)
```
:::

::: {.column width="50%"}
::: incremental
-   Important omitted variable is GDP per capita.

-   What could be the bias?

-   Does the BIAS:

    -   increases or decreases the coefficient of interest?
    -   makes the coefficient of interest significant or insignificant?
:::
:::
:::

------------------------------------------------------------------------

### Educated guess: effect of the omitted variable (2/3) {.smaller}

::: incremental
-   $\textit{OVB} = \pi_1 \times \gamma$

    -   $\pi_1^l$ is the coef. of treatment in the auxiliary of omitted on included (indep. in short);
    -   $\gamma$ is the coef. of Omitted variable in long regression;
:::

------------------------------------------------------------------------

### Educated guess: effect of the omitted variable (3/3) {.smaller}

::: incremental
-   What would be the **Effect of GDP on fertility** (coef. $\gamma$) if we had a long regression?

    -   Decreasing effect or $\gamma < 0$ ;

-   What would be the **effect of life expectancy on the GDP**?

    -   Countries with higher GDP have longer life expectancy.
    -   Although, the causal direction is not clear here, $\pi_1 > 0$

-   Therefore: $\textit{OVB} < 0$

    -   ::: {.callout-important appearance="minimal"}
        the OV **biases our estimates** by making coefficient of **life expectancy** lower than it actually is.
        :::
:::

## Calculating Omitted Variable Bias

Process:

1.  Estimate long and auxiliary regressions

2.  Extract relevant coefficients and calculating the bias

------------------------------------------------------------------------

### Step 1. Estimate long and auxiliary regressions

```{r}
#| code-fold: true
fit_l <- lm(fertility ~ lifeExpF + pctUrban + group + log(ppgdp), un_dta)
fit_ax <- lm(log(ppgdp) ~ lifeExpF + pctUrban + group , un_dta)
screen_many_regs(fit_ax, fit_l)
```

------------------------------------------------------------------------

### Step 2. Extract relevant coefficients and calculating the bias {.smaller}

```{r}
beta_s <- fit_s %>% tidy() %>% filter(term == "lifeExpF") %>% pull(estimate)
beta_l <- fit_l %>% tidy() %>% filter(term == "lifeExpF") %>% pull(estimate)
pi_1 <- fit_ax %>% tidy() %>% filter(term == "lifeExpF") %>% pull(estimate)
gam <- fit_l %>% tidy() %>% filter(term == "log(ppgdp)") %>% pull(estimate)
```

OVB:

```{r}
beta_s - beta_l
```

or OVB:

```{r}
pi_1 * gam
```

## Interpreting the results with OVB

-   OVB affects not only the variable of interest, but also all other variables in the regression.

```{r}
screen_many_regs(fit_s, fit_l)
```

# Example 2: Wage, education and ability

# References
