---
title: "AE12 Instrumental Variable"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(alr4)
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)
library(correlation)
library(patchwork)
library(lmtest)
library(sandwich)
library(clubSandwich)
library(forcats)
library(modelbased)
library(emmeans)
library(ggeffects)
library(insight)
library(scales)
library(skimr)
library(glue)

ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 10,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "100%", 
  message = FALSE,
  echo = TRUE, 
  cache = FALSE
)
```

# Goals:

-   learn how to fit the instrumental variable regression.

# Exercise 1. Reproduce theffect of birth quarter on schooling

Angrist, J. D., & Krueger, A. B. (1991). Does Compulsory School Attendance Affect Schooling and Earnings? The Quarterly Journal of Economics, 106, 979--1014. https://doi.org/10.2307/2937954

# 1 Data loading

Following variables are there:

lnw - log if wage
s - years of schooling
yob - year of birth
qob - quarter of birth
sob - week of birth


```{r}
library(haven)
dta <- 
  read_dta("data/ak91.dta") %>% 
  mutate(across(c(yob, qob, sob), ~ as.factor(.))) %>% 
  rename(schooling = s)
glimpse(dta)
skim(dta)
```

## 1.2 Data visualisation

```{r}
library(lubridate)
plot_dta <- 
  dta %>% 
  mutate(date = str_c("19", yob, "-", qob) %>% 
           lubridate::yq()) %>% 
  group_by(date) %>%
  summarise(across(c(lnw, schooling), ~mean(.))) %>% 
  mutate(yob = lubridate::year(date),
         qob = lubridate::quarter(date) %>% as_factor())
  
plot_dta %>% 
  ggplot() + 
  aes(y = lnw, x = date) + 
  geom_path() + 
  geom_label(aes(label = qob, fill = qob), colour = "white", fontface = "bold") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  xlab("Year of birth") + 
  ylab("Log of the weekly wage")

plot_dta %>% 
  ggplot() + 
  aes(y = schooling, x = date) + 
  geom_path() + 
  geom_label(aes(label = qob, fill = qob), colour = "white", fontface = "bold") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  xlab("Year of birth") + 
  ylab("Years of schooling")
```


# 2 produce IV using lm (only for the example)

Never use such manual 2sls estimations in real life as your SE are inefficient.

## 2.1 OLS

```{r}
fit1 <- lm(lnw ~ schooling + yob, dta)
summary(fit1)
```


## 2.2 first stage with the 4th qurter only

```{r}
fit_first_stage_qob1 <- lm(schooling ~ yob + qob1, 
                      dta %>% mutate(qob1 = as.integer(qob == 1)))
fit_first_stage_qob1_noiv <- lm(schooling ~ yob, 
                                dta %>% mutate(qob1 = as.integer(qob == 1)))
summary(fit_first_stage_qob1)
```


## 2.2 first stage with all quarters of birth

```{r}
fit_first_stage_allqob <- lm(schooling ~ yob + qob, dta)
summary(fit_first_stage_allqob)
```


## 2.3 second stage with the 4th qurter only

```{r}
fit_second_stage_qob1 <-
  lm(lnw ~ schooling_hat + yob, 
     dta %>% mutate(schooling_hat = fitted(fit_first_stage_qob1)))

fit_second_stage_allqob <-
  lm(lnw ~ schooling_hat + yob, 
     dta %>% mutate(schooling_hat = fitted(fit_first_stage_allqob)))

```

## 2.4 compare the models

```{r}
library(modelsummary)
modelsummary(
  list(
    `OLS` = fit1,
    `First stage without IV` = fit_first_stage_qob1_noiv, 
    `First stage QOB = 1` = fit_first_stage_qob1,
    `First stage all QOB` = fit_first_stage_allqob,
    `Second stage QOB = 1` = fit_second_stage_qob1,
    `Second stage all QOB` = fit_second_stage_allqob
  ),
  fmt = "%.3f",
  estimate = "{estimate}{stars} ({std.error})",
  statistic = NULL,
  gof_map = c(
    "adj.r.squared",
    "nobs",
    "statistic.Weak.instrument",
    "statistic.Wu.Hausman",
    "F",
    "p.value"
  ),
  coef_rename =
    c(
      "schooling" = "Years of schooling",
      "schooling_hat" = "Years of schooling"
    ),
  coef_omit = "yob|Intercept",
  stars = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  diagnostic = TRUE,
  title = "Manual 2SLS"
)
```

## 3. Make an IV estimation

```{r}
library(AER)

iv_q1 <- ivreg(
  lnw  ~ schooling + yob | qob1 + yob,
  data = dta %>% mutate(qob1 = as.integer(qob == 1))
)
summary(iv_q1, diagnostics = T)


iv_qall <- ivreg(lnw  ~ schooling + yob | qob + yob,  data = dta )
summary(iv_qall, diagnostics = T)
```

