---
title: "AE04-02 Multiple Linear Regression: linearity"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)       # for data wrangling
library(alr4)            # for the data sets #
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)

ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.width = 10,
  fig.asp = 0.618,
  out.width = "100%"
)
```

## Goals

-   Perform a linear regression analysis;
-   Learn how to test linearity assumption;
-   Practice linear transformation;
-   Practice coefficients interactions;
-   Exerciser coefficients interpretation with linear transformation and or interactions;

## Data

We explore `UN11` data from with 199 observations.

Variables are:

-   `fertility` - number of children per woman;
-   `lifeExpF` - Female life expectancy, years;
-   `ppgdp` - Per capita gross domestic product in US dollars;
-   `pctUrban` - Percent of Urban population;
-   `group` - variable with 3 values "oecd", "africa" and "others";

### Data loading

```{r}
#| code-fold: true
library(alr4)
library(tidyverse)
un_dta <- 
  alr4::UN11 %>%
  as_tibble()
glimpse(un_dta)
```


## Model

Let us build a simple econometric model: 

$$\textit{fertility} = f(\textit{ppgdp}, \textit{pctUrban})$$

## Ex. 1. Regression

Check help: `?lm`

```{r}
# ft1 <- lm(_____ ~ ______, data = _____)
```


## Ex. 1.1 Reg. Summary with `parameters` and `performance`

```{r}
# library(______)
# library(______)
# parameters(______)
# performance(______)
```

## Ex. 2. Linearity

### Ex. 2.1 Checking linearity with `base::plot`

Important to specify argument `which = 1` that makes R print only the first plot!

```{r}
# plot(______, which = 1)
```


### Ex. 2.2 Checking linearity with `performance::check_model`

See help: `?check_model`. First argument is the model, second argument `check = "linearity"` make sure that the function checks linearity only. 

```{r}
# _____(_____, check = "linearity")
```

### Ex. 2.3 Checking linearity with `car::residualPlots`

```{r}
# library(car)
# residualPlots(_____, plot = FALSE)
```

```{r}
# _____(_____, test = FALSE)
```

## Ex. 3. Linear transformation

What variable should we first try to transform with log?


### Ex. 3.1 Linear transformation of ______________

```{r}
# ft2 <- lm(_______ ~ log(______) + _______, ________)
```


### Ex. 3.2 Checking linearity with `car::residualPlots`

```{r}
# _______(_______)
```

Has anything changed? Do we have an improvement in the linearity assumption?



### Ex. 3.3 Interpretation logged variable

```{r}
# parameters(_______)
# _______(ft2)
```

### Ex. 3.4 Predicted values when regressor is transformed

`?ggeffects::ggpredict`

Firs argument of a function is the model, second argument is the term that we want to predict. 

if we pass result of `ggpredict` to plot, we will get a plot of predicted values constructed for us. 

```{r}
library(ggeffects)
# ggpredict(_______, terms = "_______")
# _______(ft2, _______) %>% plot()
```

## Ex. 4 Introducing an interaction term

### Ex. 4.1 Regression with an interaction term

```{r}
# ft3a <- lm(_____ ~ log(_____) * _____, data = un_dta)
```

We can also calculate log of the variable before regression, in the data and then use a newly calculated variable:

```{r}
# ft3 <- lm(_____ ~ log_ppgdp * _____ ,
#           data = 
#             un_dta %>%
#             mutate(log_ppgdp = log(ppgdp))
#           )
```

### Ex. 4.2 Checking linearity with `car::residualPlots`

Are there any improvement in fulfillment of the linearity assumption?

```{r}
# _______(_______)
```

### Ex. 4.3 Interpretation logged variable + interaction term

```{r}
# parameters(_______)
# _______(ft3)
```

### Ex. 4.4 Maginal effects

```{r}
# library(modelbased)
# library(emmeans)
# estimate_slopes(_______, trend = "log_ppgdp")
# estimate_slopes(_______, trend = "_______", at = "pctUrban") %>% plot()
```

```{r}
# estimate_slopes(ft3, trend = "pctUrban")
# estimate_slopes(ft3, trend = "_______", at = "_______") %>% plot()
```


### Ex. 4.5 Predicted values

Let us use regression `fit3a`

```{r}
# library(ggeffects)
# ggpredict(______, terms = "ppgdp")
# ggpredict(______, terms = c("______", "pctUrban [10, 50, 90]")) %>% plot()
```


## Solutions

```{r}
## Regression
ft1 <- lm(fertility ~ ppgdp + pctUrban, data = un_dta)

## Reg. Summary
library(parameters)
library(performance)
parameters(ft1)
performance(ft1)

## Linearity

### Checking linearity with `base::plot`
plot(ft1, which = 1)

### Checking linearity with `performance::check_model`
check_model(ft1, check = "linearity")

### Checking linearity with `car::residualPlots`
library(car)
residualPlots(ft1)

## Linear transformation of ______________
ft2 <- lm(fertility ~ log(ppgdp) + pctUrban, data = un_dta)

## Checking linearity with `car::residualPlots`
residualPlots(ft2)

## Interpretation logged variable
parameters(ft2)
performance(ft2)

## Predicted values
library(ggeffects)
ggpredict(ft2, terms = "ppgdp")
ggpredict(ft2, terms = "pctUrban ")
ggpredict(ft2, terms = "ppgdp") %>% plot()

## Introducing interaction term
ft3 <- lm(fertility ~ log_ppgdp + pctUrban + log_ppgdp:pctUrban ,
          data = un_dta %>% mutate(log_ppgdp = log(ppgdp)))

ft3a <- lm(fertility ~ log(ppgdp) * pctUrban, data = un_dta)

## Checking linearity with `car::residualPlots`
residualPlots(ft3)

## Interpretation logged variable + interaction term
parameters(ft3)
performance(ft3)

## Maginal effects
library(modelbased)
library(emmeans)
estimate_slopes(ft3, trend = "log_ppgdp")
estimate_slopes(ft3, trend = "pctUrban")

estimate_slopes(ft3, trend = "log_ppgdp", at = "pctUrban") %>% plot()
estimate_slopes(ft3, trend = "pctUrban", at = "log_ppgdp") %>% plot()

## Predicted values
library(ggeffects)
ggpredict(ft3a, terms = "ppgdp")
ggpredict(ft3a, terms = c("ppgdp", "pctUrban [10, 50, 90]")) %>%  plot()
```

