---
title: "AE04-01 Multiple Linear Regression"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---

# Setup

```{r}
library(tidyverse)
library(wooldridge)
library(modelsummary)
library(GGally)
library(parameters)
library(performance)
library(see)
library(patchwork)

ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.width = 12,
  fig.asp = 0.618,
  out.width = "100%"
)
```

## Goals

-   Fit multiple linear regression;
-   Display and interpret regression summary;
-   Compute fitted values and residuals;

## Data

We use data from (Blackburn and Neumark, 1992) on wage determinants. Variables present are:

-   $wage$ - monthly earnings in USD;
-   $educ$ - years of education;
-   $exper$ - years of experiences;
-   $black$ - dummy variable representing individuals which are not Caucasian;

See: Blackburn, M. and Neumark, D. (1992) Unobserved Ability, Efficiency Wages, and Interindustry Wage Differentials. The Quarterly Journal of Economics , Vol. 107, No. 4. Oxford University Press (OUP). p. 1421-1436

```{r}
wage_dta <- 
    wooldridge::wage2 %>% 
    as_tibble() %>% 
    select(wage, educ, exper, black)
```

## Ex. 1. Glimpse at the data

```{r}
# __________(_____)
```

Compute a new variables `caucasian` that takes values "yes" and "no" for each person depending on the values of the variable `black`.

```{r}
# wage_dta <-
#   _____ %>%
#   mutate(_____ = ifelse(black, "no", "yes"),
#          caucasian = as.factor(caucasian))
# glimpse(_____)
```

## Ex. 2. Descriptive statistics

Use function `datasummary_skim()` from the package `modelsummary`.

```{r}
# library(_________)
# _________(wage_dta)
```

## Ex. 3. Visual data inspection

We inspect data visually by building a grid of scatter plots using `GGally::ggpairs()`

```{r}
# library(_________)
# _________(wage_dta)
```

Add parameter `aes(colour = caucasian)` to `ggpairs()`.

What is the difference compare to the previous plot?

```{r}
# ggpairs(_________, _________)
```

## Ex. 4. Fit regressions

Dependent variable is `wage`, independent are: `educ`, `exper` and `black`

```{r}
# fit1 <- lm(____ ~ ____ + ____ + ____, data = ____)
# fit1
```

## Ex. 5. Print regression results

```{r}
# fit1
```

## Ex. 6. Make a summary of regression results

```{r}
# summary(______)
```

Interpret the coefficients:

## Ex. 7. Regression summary using `parameters::parameters()`

Check help on `?parameters::parameters`

```{r}
# library(________)
# ________(________)
```

## Ex. 8. Regression Goodness of fit using `performance::performance()`

Check help on `?performance::performance`

```{r}
# library(________)
# performance(________)
```

## Ex. 9. Extract fitted values to an object

```{r}
# fitted_vector <- fitted(________)
# fitted_vector[1:20]
```

## Ex. 10. Extract residuals to an object

```{r}
# resid_vector <- resid(________)
# resid_vector[1:20]
```

## Ex. 11. Make a plot of residuals vs fitted

Use function `plot()`. Put fitted as an `x` argument and residuals as `y`.

```{r}
# plot(x = _____, ___ = ________)
```

## Ex. 12. Make a plot of residuals vs fitted using `check_model` from `performance` package

See: `?performance`

```{r}
# check_model(fit1, check = "linearity")
```

## Solutions

```{r}
#| code-fold: true
library(tidyverse)
library(wooldridge)

wage_dta <- 
    wooldridge::wage2 %>% 
    as_tibble() %>% 
    select(wage, educ, exper, black)
glimpse(wage_dta)

wage_dta <-
  wage_dta %>%
  mutate(caucasian = ifelse(black, "no", "yes"),
         caucasian = as.factor(caucasian))
glimpse(wage_dta)

library(modelsummary)
datasummary_skim(wage_dta)

library(GGally)
ggpairs(wage_dta)

ggpairs(wage_dta, aes(colour = caucasian))

fit1 <- lm(wage ~ educ + exper + black, data = wage_dta)
fit1

summary(fit1)

library(parameters)
parameters(fit1)

library(performance)
performance(fit1)

fitted_vector <- fitted(fit1)
fitted_vector[1:20]

resid_vector <- resid(fit1)
resid_vector[1:20]

plot(x = fitted_vector, y = resid_vector)

check_model(fit1, check = "linearity")
```
