---
title: "AE04-02 Multiple Linear Regression: Hedonic Prices"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(alr4)
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 12,
  fig.asp = 0.618,
  fig.retina = 1,
  out.width = "100%", 
  message = FALSE,
  echo = TRUE
)

my_gof <- function(fit_obj, digits = 4) {
  sum_fit <- summary(fit_obj)
  
  stars <- 
    pf(sum_fit$fstatistic[1],
       sum_fit$fstatistic[2], 
       sum_fit$fstatistic[3],
       lower.tail=FALSE) %>% 
    symnum(corr = FALSE, na = FALSE, 
           cutpoints = c(0,  .001,.01,.05,  1),
           symbols   =  c("***","**","*"," ")) %>% 
    as.character()
  
  list(
    # `R^2` = sum_fit$r.squared %>% round(digits),
    # `Adj. R^2` = sum_fit$adj.r.squared %>% round(digits),
    # `Num. obs.` = sum_fit$residuals %>% length(),
    `Num. df` = sum_fit$df[[2]],
    `F statistic` = 
      str_c(sum_fit$fstatistic[1] %>% round(digits), " ", stars)
  )
}

# Function for screening many regressors
screen_many_regs <-
  function(fit_obj_list, ..., digits = 4, single.row = TRUE) {
    
    if (class(fit_obj_list) == "lm") 
      fit_obj_list <- list(fit_obj_list)
    
    if (length(rlang::dots_list(...)) > 0)  
      fit_obj_list <- fit_obj_list %>% append(rlang::dots_list(...))
    
    # browser()
    fit_obj_list %>%
      screenreg(
        custom.note =
          map2_chr(., seq_along(.), ~ {
            str_c("Model ", .y, " ", as.character(.x$call)[[2]])
          }) %>%
          c("*** p < 0.001; ** p < 0.01; * p < 0.05", .) %>%
          str_c(collapse = "\n") ,
        digits = digits,
        single.row = single.row,
        custom.gof.rows =
          map(., ~my_gof(.x, digits)) %>%
          transpose() %>%
          map(unlist),
        reorder.gof = c(3, 4, 5, 1, 2)
      )
  }
```

# Goals:

-   Get a brief introduction to the Hedonic Prices model;
-   Practice fitting linear regression in R;
-   Test linearity assumption: visually and using a statistical tests;
-   Practice linear transformation;
-   Exerciser coefficients interpretation;

# Exercise

## Data loading

```{r}
dta <- 
  alr4::MinnLand %>% 
  as_tibble()  %>%
  # filter(year == 2007) %>% 
  select(acrePrice, acres, region, year, 
         tillable, crpPct, productivity)
```

Glimpse at the data:

```{r}
# _______(dta)
```

## Summary Statistics

This is a simple short-cut to the summary statistics.

```{r}
library(modelsummary)
#dta %>% datasummary_skim()
```

### Do yourself at home!

In the code chunk below, adopt the R code used to prepare data summary manually. Run it step by step and comment on what it does between each component of the pipe.

```{r}
# dta %>%
#   _______(id = _______()) %>% 
#   _______(
#     cols = c(acrePrice, acres, year, tillable, crpPct, productivity),
#     names_to = "var",
#     values_to = "val"
#   ) %>% 
#   _______(var) %>% 
#   _______(_______(
#     c(val),
#     _______(
#       mean = ~ mean(.x, na.rm = TRUE),
#       sd = ~ sd(.x, na.rm = TRUE),
#       meadian = ~ median(.x, na.rm = TRUE),
#       n_miss = ~ sum(is.na(.x), na.rm = TRUE),
#       min = ~ min(.x, na.rm = TRUE),
#       max = ~ max(.x, na.rm = TRUE)
#     )
#   ), 
#   n = n())
```

## Visual inspection

We reply on the visual inspection of data to build some prior expectations. We will use function `ggpairs()` from `GGally`. Insert function name where appropriate:

```{r}
library(GGally)
# dta %>%
#   select(acrePrice, acres, tillable, 
#          crpPct, productivity) %>%
#   ________()
```

## Regression 1

Let us fit the basic regression and summarize the results.

```{r}
# fit1 <- lm(
#   acrePrice ~ _____ + _____ + _____ + _____ + ____ + _____,
#   data = dta
# )
# summary(fit1)
```

Use `performance` and `parameters` package to summarize the regression results.

-   Provide object with the fitted regression;

```{r}
library(parameters)
library(performance)
# parameters(_____)
# performance(_____)
```

### Linearity: visual inspection

We can use built-in function for plotting residuals vs fitted.

-   Provide object with the fitted regression;

```{r}
# plot(_____, which = 1)
```

We can also rely on the `performance` package:

```{r}
library(see)
# check_model(_____, check = "linearity", panel = FALSE)
```

What do you conclude about the linearity assumption based on these results?

### Linearity: statistical test

What does the `residualPlots()` test and plots show us?

```{r}
library(car)
# residualPlots(_____)
```

What we conclude based on this linearity test?

## Regression 2

Let us transform the dependent variable with a logarithm!

```{r}
# fit2 <- lm(
#   log(acrePrice) ~ _______________,
#   data = _____
# )
# summary(_____________)
```

### Regressions comparison

Let us compare regression 1 and 2 using function `screen_many_regs`.

```{r}
# screen_many_regs(_______, fit2)
```

### Linearity assumption

Is the linearity assumption fulfilled now?

```{r}
# residualPlots(___________)
```

# Solutions

```{r}
#| code-fold: true
dta <- 
  alr4::MinnLand %>% 
  as_tibble()  %>%
  # filter(year == 2007) %>% 
  select(acrePrice, acres, region, year, 
         tillable, crpPct, productivity)
glimpse(dta)

# Summary statistics:
dta %>%
  mutate(id = row_number()) %>% 
  pivot_longer(
    cols = c(acrePrice, acres, year, tillable, crpPct, productivity),
    names_to = "var",
    values_to = "val"
  ) %>% 
  group_by(var) %>% 
  summarise(across(
    c(val),
    list(
      mean = ~ mean(.x, na.rm = TRUE),
      sd = ~ sd(.x, na.rm = TRUE),
      meadian = ~ median(.x, na.rm = TRUE),
      n_miss = ~ sum(is.na(.x), na.rm = TRUE),
      min = ~ min(.x, na.rm = TRUE),
      max = ~ max(.x, na.rm = TRUE)
    )
  ), 
  n = n())

# Summary statistics lazy
dta %>% datasummary_skim(output = "markdown")

# Visual data inspection
dta %>%
  select(acrePrice, acres, tillable, 
         crpPct, productivity) %>%
  ggpairs()

# Regression
fit1 <- lm(
  acrePrice ~ crpPct + acres + region + 
    year + tillable + productivity, 
  data = dta
)
summary(fit1)

# Linearity
plot(fit1, which = 1)

# Linearity 2
residualPlots(fit1)

# Regression 2
fit2 <- lm(
  log(acrePrice) ~  crpPct + acres + region + 
    year + tillable + productivity,
  data = dta
)
summary(fit2)

# Comparing regressions
screen_many_regs(fit1, fit2)

# Comparing regressions
residualPlots(fit2)
```
