---
title: "AE04-01 Multiple Linear Regression"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(modelsummary)
library(GGally)
library(parameters)
library(performance)
library(see)
library(patchwork)
library(ggeffects)

ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.width = 12,
  fig.asp = 0.618,
  out.width = "100%"
)
```

## Goals

-   Fit multiple linear regression;
-   Display and interpret regression summary;
-   Compute fitted values and residuals;

## Data

We use data from (Blackburn and Neumark, 1992) on wage determinants. Variables present are:

-   $wage$ - monthly earnings in USD;
-   $educ$ - years of education;
-   $exper$ - years of experiences;
-   $black$ - dummy variable representing individuals which are not Caucasian;

See: Blackburn, M. and Neumark, D. (1992) Unobserved Ability, Efficiency Wages, and Interindustry Wage Differentials. The Quarterly Journal of Economics , Vol. 107, No. 4. Oxford University Press (OUP). p. 1421-1436

```{r}
wage_dta <- 
    wooldridge::wage2 %>% 
    as_tibble() %>% 
    select(wage, educ, exper, black)
```

## Ex. 1. Glimpse at the data

```{r}
# __________(_____)
```

Compute a new variables `caucasian` that takes values "yes" and "no" for each person depending on the values of the variable `black`.

```{r}
# wage_dta <-
#   _____ %>%
#   mutate(_____ = ifelse(black, "no", "yes"),
#          caucasian = as.factor(caucasian))
# glimpse(_____)
```

### Ex. 1.2. Descriptive statistics

Use function `datasummary_skim()` from the package `modelsummary`.

```{r}
# library(_________)
# _________(wage_dta)
```

### Ex. 2.3. Visual data inspection

We inspect data visually by building a grid of scatter plots using `GGally::ggpairs()`

```{r}
# library(_________)
# _________(wage_dta)
```

Add parameter `aes(colour = caucasian)` to `ggpairs()`.

What is the difference compare to the previous plot?

```{r}
# ggpairs(_________, _________)
```

## Ex. 2. Fit regressions

Dependent variable is `wage`, independent are: `educ`, `exper` and `black`

```{r}
# fit1 <- lm(____ ~ ____ + ____ + ____, data = ____)
# fit1
```

### Ex. 2.1. Print regression results

```{r}
# fit1
```

## Ex. 3 Regression summary

```{r}
# summary(______)
```

Interpret the coefficients:

### Ex. 2.1 Regression summary using `parameters::parameters()`

Check help on `?parameters::parameters`

```{r}
# library(________)
# ________(________)
```

### Ex. 2.2. Goodness of fit using `performance::performance()`

Check help on `?performance::performance`

```{r}
# library(________)
# performance(________)
```

## Ex. 4. Fitted

```{r}
# fitted_vector <- fitted(________)
# fitted_vector[1:20]
```

## Ex. 5. Residuals

```{r}
# resid_vector <- resid(________)
# resid_vector[1:20]
```

## Ex. 6. Residuals vs fitted plot

Use function `plot()`. Put fitted as an `x` argument and residuals as `y`.

```{r}
# plot(x = _____, ___ = ________)
```

### Ex. 6.1 Residuals vs fitted using `check_model` from `performance` package

See: `?performance::check_model`

```{r}
# check_model(fit1, check = "linearity")
```

## Ex. 7 Predicting values using regression

We can compute predicted values based on the regression results for arbitrary values of $X$. See: `?predict`

First, we need to make a table the $X$ variables for which we want compute the predicted value.

-   Variables names in table should match the variables in the regression.

Let us predict wage for a person with 0 years of education, 0 experience, when he/she is white.

```{r}
# pred1 <-
#   tibble(educ = 0,
#          exper = 0,
#          black = 0)
# predict(fit1, pred1)
```

The same when the person is black.

```{r}
# pred2 <-
#   tibble(educ = c(0, 0),
#          exper = c(0, 0),
#          black = c(0, 1)
#   )
# predict(fit1, pred2)
```

### Ex 7.1 Writing predicted values to a data frame

Use mutate to mutate new variable in a data frame:

```{r}
# pred3 <-
#   tibble(educ = c(0, 0, 10),
#          exper = c(0, 0, 25),
#          black = c(0, 1, 1)
#   )
# 
# pred3 %>% 
#   mutate(predicted = predict(fit1, pred3))
```

### Ex 7.2 Visual inspection of predicted values

We can use function `ggeffects::ggpredict()`, check help.

```{r}
# library(ggeffects)
# ggpredict(fit1, term = "educ")
```

We can also plot this effects using `plot` after `ggpredict`


```{r}
# ggpredict(fit1, term = "educ") %>% ________()
```

Compute predicted values for a different independent variable and a dummy variable

```{r}
# ggpredict(fit1, term = c("________", "________")) 
# ________(________, term = c("________", "________")) %>% plot()
```

## Solutions

```{r}
#| code-fold: true
library(tidyverse)
library(wooldridge)

wage_dta <- 
    wooldridge::wage2 %>% 
    as_tibble() %>% 
    select(wage, educ, exper, black)
glimpse(wage_dta)

wage_dta <-
  wage_dta %>%
  mutate(caucasian = ifelse(black, "no", "yes"),
         caucasian = as.factor(caucasian))
glimpse(wage_dta)

library(modelsummary)
datasummary_skim(wage_dta)

library(GGally)
ggpairs(wage_dta)

ggpairs(wage_dta, aes(colour = caucasian))

fit1 <- lm(wage ~ educ + exper + black, data = wage_dta)
fit1

summary(fit1)

library(parameters)
parameters(fit1)

library(performance)
performance(fit1)

fitted_vector <- fitted(fit1)
fitted_vector[1:20]

resid_vector <- resid(fit1)
resid_vector[1:20]

plot(x = fitted_vector, y = resid_vector)

check_model(fit1, check = "linearity")

pred1 <-
 tibble(educ = 0,
        exper = 0,
        black = 0)
predict(fit1, pred1)

pred2 <-
  tibble(educ = c(0, 0),
         exper = c(0, 0),
         black = c(0, 1)
  )
predict(fit1, pred2)

pred3 <-
  tibble(educ = c(0, 0, 10),
         exper = c(0, 0, 25),
         black = c(0, 1, 1)
  )
pred3 %>% 
  mutate(predicted = predict(fit1, pred3))

library(ggeffects)

ggpredict(fit1, term = "educ")
ggpredict(fit1, term = "educ") %>% plot()

ggpredict(fit1, term = c("exper", "black"))
ggpredict(fit1, term = c("exper", "black")) %>% plot()
```
