---
title: "AE11-01 translog production function"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(alr4)
library(GGally)
library(parameters)
library(performance)
library(see)
library(car)
library(broom)
library(modelsummary)
library(texreg)
library(correlation)
library(patchwork)
library(lmtest)
library(sandwich)
library(clubSandwich)
library(forcats)
library(modelbased)
library(emmeans)
library(ggeffects)
library(insight)
library(scales)
library(glue)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 16,
  fig.asp = 0.618,
  fig.retina = 1,
  out.width = "100%", 
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)


ggplot2::theme_set(ggplot2::theme_bw())

knitr::opts_chunk$set(
  fig.width = 10,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "100%", 
  message = FALSE,
  echo = TRUE, 
  cache = TRUE
)

# Custom functions to summaries data nicely
get_signif <- 
  function(x) {
    symnum(
      x,
      corr = FALSE,
      na = FALSE,
      cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
      symbols = c("***", "**", "*", ".", " ")
    ) %>% 
      as.character()
  }

tidy_skim <- 
  function(dta) {
    dta %>%
      select(- any_of(c("id", "time"))) %>% 
      skimr::skim_without_charts() %>%
      as_tibble() %>%
      select(any_of(c("skim_variable","n_missing")), contains("numeric")) %>%
      rename_with( ~ str_remove(., "numeric\\."))
  }

tidy_coeftest <- 
  function(
    mod, 
    mod_name = deparse(substitute(mod)), 
    mod_vcov = vcov(mod),
    dig = 3, 
    ...) {
    mod_name_sym <- sym(mod_name)
    mod %>%
      lmtest::coeftest(vcov. = mod_vcov)  %>%
      broom::tidy() %>%
      mutate(
          across(c(estimate, std.error),
                 ~ scales::number(., 1 / 10 ^ dig, big.mark = ",")),
        across(c(p.value), ~ insight::format_p(., stars_only = TRUE)),
        mod_stat := glue::glue("{estimate}{p.value} ({std.error})")
      ) %>%
      select(parameter = term, !!mod_name_sym := mod_stat)
  }

tidy_gof <- 
  function(
    mod, 
    mod_name = deparse(substitute(mod)), 
    dig = 3, 
    ...) {
    mod_sum <- summary(mod)
    mod_sum <- mod_sum$fstatistic
    if (is.vector(mod_sum)) {
      df1 <- mod_sum[[2]]
      df2 <- mod_sum[[3]]
      df <- str_c(c(df1, df2), collapse = "; ")
    } else {
      df <- str_c(mod_sum$parameter, collapse = "; ")
    }
    mod %>%
      broom::glance() %>%
      {
        dta <- .
        if (!"logLik" %in% names(dta)) {
          dta <-
            mutate(dta, logLik = mod %>% stats::logLik() %>% as.numeric())
        }
        
        if (!"AIC" %in% names(dta)) {
          dta <- mutate(dta, AIC = mod %>% stats::AIC() %>% as.numeric())
        }
        
        if (!"BIC" %in% names(dta)) {
          dta <- mutate(dta, BIC = mod %>% stats::BIC() %>% as.numeric())
        }
        dta
      } %>%
      mutate(
        across(any_of(c("r.squared", "deviance", "adj.r.squared")), 
               ~ scales::number(., 1 / 10 ^ dig, big.mark = ",")),
        across(any_of(c("statistic", "logLik", "AIC", "BIC")),
               ~ scales::number(., 1, big.mark = ",")),
        `F Statistics (df)` =
          glue("{statistic}{get_signif(p.value)} ", "({df})"),
        nobs = scales::number(nobs, 1, big.mark = ",")
      ) %>%
      select(
        N = nobs,
        `R-sq. adj.` = adj.r.squared,
        `Log likelihood` = logLik,
        AIC,
        BIC,
        `F Statistics (df)`
      ) %>%
      pivot_longer(everything(), 
                   names_to = "parameter", 
                   values_to = mod_name)
  }


tidy_summary <-
  function(mod,
           mod_name = deparse(substitute(mod)),
           mod_vcov = vcov(mod),
           dig = 3,
           ...) {
    
    tidy_coeftest(mod,mod_name = mod_name, mod_vcov = mod_vcov, dig = dig) %>% 
      bind_rows(tidy_gof(mod, mod_name = mod_name, dig = dig))
  }


tidy_summary_list <-
  function(mod_list,
           mod_vcov = NULL,
           dig = 3,
           ...) {
    # browser()
    mod_list %>%
      list(., names(.), seq_along(.)) %>% 
      pmap(~ {
        vcov_here <- vcov(..1)
        if (!is.null(mod_vcov[[..3]]))
          vcov_here <- mod_vcov[[..3]]
        tidy_summary(
          mod = .x,
          mod_name = .y,
          mod_vcov = vcov_here,
          dig = dig
        )
      }) %>%
      reduce(full_join, by = "parameter")
  }
```

# Goals:

-   Learn what is the production function and how to fit it using panel data framework.

# Exercise 1. Effect of labour union on the firms value-added peoduced

Using the data for Ghanaian firms -- `Ghana_Firms_JDE04` -- estimate the following production function by OLS, FE, FD and RE: $y_{it} = \alpha + \beta_n n_{it} + \beta_k k_{it} + \gamma \cdot union$,

where:

-   $y_{it}$ is the natural logarithm of value-added of a firm;

-   $n_{it}$ is log employment;

-   $k_{it}$ is log capital stock;

-   $union$ is a dummy variable equal to one if the firm is unionised and zero if it is not;

-   $i$ is a firm-specific, time invariant and unobserved effect;

-   $t$ is the time;

-   $\alpha$ is the intercept and the $\beta$ , $\gamma$ are parameters.

Following variables are present in the data:

-   `firm` is the farm ID;
-   `wave` is the time period of the data collection;
-   `lrvadid` is the log of real value-added;
-   `ll` is the log of labour input and;
-   `lk` is the log of capital;
-   `eduwgt` is a weighted average of the level of education in the firm;
-   `agewgt` is a weighted average of the age of the workers (to capture the level of general labour-market experience) and;
-   `tenwgt` is a weighted average of the tenure of workers in the firm which is intended to capture firm-specific skills;
-   `food2` is the food sector;
-   `tex_gar` is the textile-garment sector;
-   `wood` is the wood sector;
-   `furn` is the furniture sector and;
-   `metal1` is the metal sector;
-   `anyfor` is any foreign ownership;
-   `statgh` is state and Ghanaian ownership and;
-   `ghown` is exclusively Ghanaian ownership;
-   `accra` is a dummy variable for Accra the capital city;
-   `kum` for Kumasi;
-   `cape` for Cape Cost and;
-   `tak` for Takoradi;

## 1.1 Data loading

Check help for `RiceFarms`.

```{r}
library(readr)
library(haven)
library(plm)

# dta <- read_dta("data/Ghana_Firms_JDE04.dta")
# dta_p <- dta %>% pdata.frame(index = c("firm", "wave"))
# 
# # Making panel structure
# pdim(dta_p)
```

## 1.2 Estimate a production function using various panel regression methods

Select an optimal regression model explain the differences between the estimates.

```{r}

```

## 1.3 Extract total factor productivity measures from pooled ols and fe models

```{r}

```

## 1.4 Plot TFP measures agains firm size in tems of labout or capital.

-   Perform non-parametric smoothing.

```{r}

```

# Exercise 2. Estimate a translog production function

Translog function:

$$\ln y = \beta_0 + \sum_{n = 1}^{N}  \beta_n \ln x_n + \\ \frac{1}{2} \sum_{n = 1}^{N} \sum_{m = 1}^{M}  \beta_{nm} \ln x_n \ln x_m + \sum_{k = 1}^{K} \gamma_k \delta_k + \epsilon$$

where,

-   $\ln x_n \ln x_m$ are the interaction terms between all combination of two regressors.
-   Everything else is the same as in the Cobb-Douglas.

```{r}

```
