---
title: "AE03-01 Import, cleaning, summary plot"
author: "Eduard Bukin"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
bibliography: ../../references.bib
---

## Setups

We will us `tidyverse` for data wrangling and `readr` and `readxl` for data import.

```{r setup, message=FALSE}
library(tidyverse)
library(readr)       # install.packages("readr")
library(readxl)      # install.packages("readxl")
library(janitor)     # install.packages("janitor")
library(skimr)       # install.packages("skimr")
```

## Data import and cleaning

In this exercise, we shall practice the following:

-   Loading data: `readr::read_csv()` and `readxl::excel_sheets()` with `readxl::read_excel()`;
-   Inspecting data: `dplyr::glimpse()`, `utils::View()`, `utils::head()`, `base::names()`;
-   Data summary: `base::summary()`;
-   Renaming variables: `janitor::clean_names(...)`, `dplyr::rename()`;
-   Removing undesired observations/row by their number in the data with `dplyr::slice()`;
-   Mutating/modifying types of existing variables: `dplyr::mutate()`
-   Converting excel dates to R-relevant `<date>` variable type: `janitor::convert_to_date()` and `janitor::excel_numeric_to_date()`.
-   Summary statistics with: `skimr::skim()`.
-   Plotting time series with: `ggplot2::ggplot()`, `ggplot2::aes()`, `ggplot2::geom_path()`.

## Exercise 1. Load clean and plot data from Excel

Here, we will explore the data from the file called `commodity-prices.xlsx` in the `data` folder.

### Ex. 1.1 Inspect the file in excel: `readxl::excel_sheets()`

Open the file in Excel and inspect it.

-   What sheets are there?
-   Where is the data?

Use function `excel_sheets()` to check what sheets does R sees in this file. Please note that you need to provide path to the file manually. Thus:

1.  Create an object `path_prices`

    -   and assign to it a value of the string with the path to the file.
    -   This string may look like: `./path_to_folder/file_name.ext`.
    -   Make sure that you specify the path exactly as it is with the file extension.
    -   Make sure that in the path, there are no extra spaces of characters which are not present in the file path.

2.  Execute function `excel_sheets()` specifying in the arguments path to the file.

```{r}
library(readxl)
# path_prices <- "______"
# excel_sheets(path = _______)
```

What does the `excel_sheets()` tells us?

### Ex. 1.2 Load data from the sheet: `readxl::read_excel(..., sheet = ____)`, `utils::head()`

You need to use function `read_excel()` to load data.

-   Check help for this function in the console!
-   Save data in the `environment` under the object name `prices_dta`.
-   `glimpse()` at the data
-   us function `head()` with the data and explain what it does

```{r}
# prices_dta <- 
#   ________(path_prices)

# prices_dta %>% 
#   ________()

# prices_dta %>% 
#   head()
```

As you can see, the data is loaded into R, but there are problems:

-   Variables names are long, bulky, contain spaces and it is difficult to use them.
-   All variables are in `<chr>`, which means character, when it should be numerical `<dbl>`.
-   First row in the data contains text, which is irrelevant to the data.
-   The data is not tidy! We need to clean it.

### Ex. 1.3 Cleaning variables names: `janitor::clean_names(...)` , `base::names()`

Currently variables names are very long.

-   Note, we use `names()` function to check variables names.

```{r}
# prices_dta %>% names()
```

We can simplify these names and make them machine readable and usable using function `janitor::clean_names()`.

-   check help for this function `?clean_names`

-   run it on the data and check the variables names

```{r}
# prices_dta %>% 
#   __________() %>% 
#   names()
```

Are those variables names useful now?

### Ex. 1.4 Cleaning variables names: `dplyr::rename()`

Well, no, these variables names are not useful. We still need to `rename()` them into something shorter. For this, we use `dplyr::rename()` . See help [here](https://dplyr.tidyverse.org/reference/rename.html).

The logit is:

-   we supply data into `rename()` with pipe: `data %>% rename()`.
-   in the `rename()` , we specify what should be the new name on the left hand side and old name on the right hand
    -   `data %>% rename(new_name = OLD_NAME)`.
-   we do not use any quotation marks.
-   Remember to repeat the data cleaning step with `janitor::lean_names()`.

Here is the example with the longest variable:

```{r}
# prices_dta %>% 
#   clean_names() %>% 
#   rename(wheat = soft_red_winter_wheat_no_2_f_o_b_us_gulf_usd_per_mt) %>% 
#   names()
```

Follow this example to rename variables into `date`, `oil`, `maize` and `urea`.

```{r}
# prices_dta %>%
#   clean_names() %>% 
#   rename(wheat = soft_red_winter_wheat_no_2_f_o_b_us_gulf_usd_per_mt, 
#          maize = ______________, 
#          date = ______________, 
#          ______________ = ______________, 
#          ______________ = ______________) %>%
#   names()
```

Did it work?

### Ex. 1.5 Removing first row with irrelevant data `dplyr::slice()`

Use function `slice(prices_dta, -1)`. Check help for `dplyr::slice()`.

-   with argument `-1` we are telling to R to drop row with the number `1`;
-   Note, wen we use pipeline `%>%` we do not need to specify data frame name `prices_dta` within the brackets `(…)` of the `slice()` call!
-   Remember to copy all the code from before, where you renamed the variables.
-   Assign value of this long pipe to the new object named `prices_dta_1`. This object will stand for the intermediary step in the data cleaning process.

```{r}
# prices_dta_1 <- 
#   prices_dta %>%
#   ### Place here the R code with renaming ###
#   ### Place here the R code with renaming ###
#   ### Place here the R code with renaming ###
#   slice(___) %>% 
#   glimpse()
```

If everything alright with the data now? Let us use `summary()` to summaries the variables and extract some numerical features:

```{r}
# prices_dta_1 %>%
#   summary()
```

Are these results meaningful?

### Ex. 1.6 Mutating variables types: `dplyr::mutate()`, `base::as.numeric()`.

As we say on the previous step, all variables have the type `character` or `<chr>`. As this is text, it is not possible to make summary statistics out of it. `R` simply does not understand that we want to used those variables as numbers as we need to explain this to `R`.

We need to:

-   `mutate()` existing variables (see definition of mutate [here](https://dictionary.cambridge.org/dictionary/english/mutate))

-   the help on the `mutate()` function is [here](https://dplyr.tidyverse.org/reference/mutate.html) with more [case example here](https://dplyr.tidyverse.org/articles/colwise.html) and a dedicated [Chapter 5.5 in R4DS](https://dplyr.tidyverse.org/articles/colwise.html);

-   and convert them to `numeric` or `double` type referred as `<dbl>`

-   to convert variable type, we use `as.numeric()`, see: `?as.numeric` or run `as.numeric(c("-.1"," 2.7 ","B"))` in console.

Here is the example that should work for one column:

-   Note, it will only work if you made previous cleaning steps correctly.

-   please un-commencement it.

```{r}
# prices_dta_1 %>% 
#   mutate(oil = as.numeric(oil)) %>% 
#   glimpse()
```

As you can see now, `oil` variable has numeric format. Thus, when we run `summary()` we get description of the numerical values there:

```{r}
# prices_dta_1 %>% 
#   mutate(oil = as.numeric(oil)) %>% 
#   summary()
```

Now, your turn to mutate all variable as numeric.

```{r}
# prices_dta_1 %>%
#   mutate(oil = as.numeric(oil),
#          wheat = ____________(____),
#          _____ = ____________(urea),
#          _____ = ____________(____),
#          _____ = ____________(____)) %>%
#   glimpse()
```

### Ex. 1.7 Mutating date variable: `dplyr::mutate()`, `janitor::convert_to_date()`.

Check results of the previous chunk!

-   It is clear that `date` variable is not a date, but a series of a number instead.
-   This is because Excel stores date as a number of days since January 1, 1990. Thus, 33664 is March 1, 1992.
-   We need to convert such date notations to some real dates using function `janitor::convert_to_date()`.
-   See: `?convert_to_date`.
-   Convert date to the type `<date>` in the same way as variable type conversion at the previous step.
-   Assign new object `prices_dta_clean` with the value of the data frame with all cleaning steps.

```{r}
# prices_dta_clean <- 
#   prices_dta_1 %>%
#   mutate(oil = as.numeric(oil),
#          wheat = ____________(____),
#          _____ = ____________(urea),
#          _____ = ____________(____),
#          date = convert_to_date(____)) 
# prices_dta_clean %>% glimpse()
```

We have manage to clean the data!

Do data `summary()` of the data to see what the variables are about.

```{r}
# 

```

### Ex. 1.8 Use `skimr::skim()` to generate summary statistics of the data

Go to [skimr website](https://docs.ropensci.org/skimr/) and learn how to use function `skim()`. Apply it to the `prices_dta_clean` data frame below and discuss how the results are different from `summary()`.

```{r}
library(skimr)       # install.packages("skimr")
# prices_dta_clean %>% 
#   _____()
```

### Ex. 1.9 Plot a time-series of all four variables using `ggplot2` package

Plotting according to the `grammar of graphics` (`gg`) using package `ggplot2` is a rewarding process. But, we need to follow some steps.

1.  We need to make data clean and tidy.
    -   Ideally, data has to be in the long format, but wide data may also work
2.  Send data to `ggplot()` function to initiate a plot.
    -   Note, we use `%>%` (pipe) for this step: `data %>% ggplot()`
3.  Start adding (`+`) various `aesthetics` to the plot using `aes()` function:
    -   Note, inside the ggplot builder we use `+`, not a pipe! `data %>% ggplot() + aes(x, y)`
4.  Add geometries using `+`
    -   `data %>% ggplot() + aes(x, y) + geom_line()`

Let us give it a try! Let us convert existing data to the plot basis and add an aesthetics for `x` axis:

-   Use `date` as an aesthetics for `x` axis, `wheat` for `y` axis and a string `"wheat"` for `color`.
-   Do not use quotation marks in `aes` for `x` and `y`.
-   Do USE quotation marks in `aes` for `colour`.

```{r}
# prices_dta_clean %>% 
#   _________() + 
#   aes(____ = date, y = _____, color = "_____") 
```

R created a grid for a plot, but no plot. This is because, we did not add any geometries! Let us add a `geom_path()` to the plot.

```{r}
# prices_dta_clean %>% 
#   _________() + 
#   aes(____ = date, y = _____, color = "_____") 
#   _________() + 
#   geom_path(aes(y = _______, colour = "Maize")) + 
#   geom_path(aes(y = _______, colour = "_____"))
```

Now, we shall add another line to the plot with the Maize prices:

-   we need to add a line with a new `geom_path()`, and
-   specify `aes()` inside `geom_path()`,
-   for example `geom_path(aes(y = maize, colour = "Maize"))`.
-   do not forget `+`.

Go ahead:

```{r}
# prices_dta_clean %>% 
#   _________() + 
#   aes(____ = date, y = _____, color = "_____") 
#   _________() + 
#   geom_path(_____(y = ______, colour = "_____"))
```

Add another line with \`geom_path\`, this time adding the oil prices to the plot.

```{r}
# prices_dta_clean %>% 
#   _________() + 
#   aes(____ = date, y = _____, color = "_____") 
#   _________() + 
#   geom_path(_____(y = ______, colour = "_____")) + 
#   geom_path(_____(y = ______, colour = "_____"))
```

### Solution to the exercise 2

```{r class.source = 'fold-hide'}
prices_dta_clean <- 
  read_excel("data/commodity-prices.xlsx", sheet =  "data") %>% 
  clean_names() %>% 
  rename(wheat = soft_red_winter_wheat_no_2_f_o_b_us_gulf_usd_per_mt, 
        maize = yellow_maize_no_2_f_o_b_us_gulf_usd_per_mt, 
        date = day_month_year, 
        oil = crude_oil_brent_usd_per_barrel, 
        urea = urea_f_o_b_black_sea_usd_per_mt) %>%
  slice(-1) %>% 
  mutate(
    oil = as.numeric(oil),
    wheat = as.numeric(wheat),
    maize = as.numeric(maize),
    urea = as.numeric(urea),
    date = convert_to_date(date)
  ) 

glimpse(prices_dta_clean)

summary(prices_dta_clean)

skim(prices_dta_clean)

prices_dta_clean %>% 
  ggplot() + 
  aes(x = date, y = wheat, colour = "Wheat") + 
  geom_path() + 
  geom_path(aes(y = maize, colour = "Maize")) + 
  geom_path(aes(y = urea, colour = "Urea"))
```

Same plot could be build build in a more simple way if we use long formatted data.

```{r class.source = 'fold-hide'}
prices_dta_clean_long <- 
  prices_dta_clean %>% 
  pivot_longer(cols = c(oil:urea))

glimpse(prices_dta_clean_long)

prices_dta_clean_long %>% 
  ggplot() + 
  aes(x = date, y = value, colour = name) + 
  geom_path()
```

## Exercise 2. OPTIONAL Import from a `.csv`: coma separated file

Here is the example [NHIS 2009](https://raw.githubusercontent.com/EBukin/datasets4examples/main/datasets-raw/NHIS2009.csv) data used in the [@Angrist2014]. To load such data in R, we can use `readr` package from tidyverse [readr](https://readr.tidyverse.org/).

One may use a visual import tool for data available in R studio. But remember to save the R code for data import in the script. Data import code must be a part of your analysis.

The challenge is to specify the right path to the file that we want to import. For this exercise, this file is also saved in the folder `./data/NHIS2009.csv`

### 2.1 Simple CSV file: `readr::read_csv()`, `dplyr::glimpse()` , `base::summary()` , `utils::View()`

In the folder `data` there is a file `chicken.csv`. Load it into the object chick.

```{r}
library(readr)
# chick <- read_csv(____)
```

Now inspect the data:

-   use `glimpse(____)`;
-   print the data set;
-   `View()` the data;

```{r}
#

```

Try function `summary(_____)`.

```{r}
# 

```

What is the difference between summary and previous two ways of data exploration?

### 2.2 Large and complex CSV file

Loading larger CSV files is not different from loading small files. Below, load the file `NHIS2009.csv` from the folder `data` and then glimpse at it:

```{r}
# 

```

### Solutions

```{r class.source = 'fold-hide'}
chick <- read_csv("data/chicken.csv")
glimpse(chick)
# View(chick)
summary(chick)
nhis <- read_csv("data/NHIS2009.csv")
```

## Exercise 3. OPTIONAL. `janitor` for cleaning data in R

we follow the data cleaning exercise form the [janitor webpage](https://github.com/sfirke/janitor). Data set is called `dirty_data.xlsx` and it is located in: `./data/dirty_data.xlsx` or on the [github](https://github.com/sfirke/janitor/raw/main/dirty_data.xlsx).

```{r}
ddta <- read_excel("./data/dirty_data.xlsx")
glimpse(ddta)
```

As you can see, this data is dirty. The problems are:

-   Variables names are wrong, they are in fact in the first row of data.
-   All variables are in character type, when in fact some variables are in different format.
-   There are some columns that are empty.
-   There are some rows that are empty.

### 3.1 Make all variables named as values in the first row.

Use function `row_to_names()` and specify the parameter `row_number` to 1.

```{r}
# ddta %>% 
#   ____________(__________ = 1) %>% 
#   glimpse()
```

### 3.2 Make all names nice

Currently, all names are (insert your code form previous step):

```{r}
# ddta %>% 
#   ____________(__________ = 1) %>% 
#   names()
```

Use function `clean_names()`, to make names nice.

```{r}
# ddta %>%
#   ____________(__________ = 1) %>%
#   ____________() %>% 
#   glimpse()
```

### 3.3 Make variable `hire_data` as date

To convert variable types to meaningful numeric variables, we need to use some more functions. Specifically, functions `convert_to_date()`. In addition, we need to use function `mutate()` to tell r to modify existing variable.

Here is the example, where we use mutate to modify existing variable `Sepal.Width` with the same variable but as character string instead of a number.

```{r}
iris %>% glimpse()

iris %>% 
  mutate(Sepal.Width = as.character(Sepal.Width)) %>% 
  glimpse()
```

Use the example from above to modify the variable `hire_date` with its values converted to date using function `convert_to_date`.

```{r}
#ddta %>%
#   ____________(__________ = 1) %>%
#   ____________() %>% 
#   mutate(_________ = ____________(hire_date)) %>% 
#  glimpse()
```

### 3.4 Make variable `percent_allocated` as numeric

Same as above, but now using function `as.numeric()`.

```{r}
# ddta %>%
#   ______________(__________ = 1) %>%
#   ______________() %>%
#   mutate(_________ = ____________(hire_date),
#          percent_allocated = ________(___________)) %>%
#   glimpse()
```

### Solutions

```{r class.source = 'fold-hide'}
ddta <- read_excel("./data/dirty_data.xlsx")
ddta %>%
  row_to_names(row_number = 1) %>%
  clean_names() %>%
  mutate(hire_date = as.numeric(hire_date),
         percent_allocated = as.numeric(percent_allocated)) %>%
  glimpse()
```
