---
title: "AE03-02 Exploring data on plstic waste"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---


This exercise is adopted from [Data Science in a Box](https://datasciencebox.org/exploring-data.html?q=import#visualising-data).

# Learning goals

-   Visualizing numerical and categorical data and interpreting visualizations
-   Recreating visualizations
-   Getting more practice using with R, RStudio, Git, and GitHub

## Setup

```{r load-packages, message=FALSE, eval=TRUE}
library(tidyverse)
```

## Data

The data set for this assignment can be found as a csv file in the `data` folder of your repository. You can read it in using the following.

```{r load-data, message=FALSE, eval=TRUE}
plastic_waste <- read_csv("data/plastic-waste.csv")
```

The variable descriptions are as follows:

-   `code`: 3 Letter country code
-   `entity`: Country name
-   `continent`: Continent name
-   `year`: Year
-   `gdp_per_cap`: GDP per capita constant 2011 international \$, rate
-   `plastic_waste_per_cap`: Amount of plastic waste per capita in kg/day
-   `mismanaged_plastic_waste_per_cap`: Amount of mismanaged plastic waste per capita in kg/day
-   `mismanaged_plastic_waste`: Tonnes of mismanaged plastic waste
-   `coastal_pop`: Number of individuals living on/near coast
-   `total_pop`: Total population according to Gapminder

# Exercises 1.

Glimpse at the data

```{r}
plastic_waste %>% glimpse()
plastic_waste %>% head()
```

# Exercises 2.

Compute summary statistics using `skimr::skim()` . Load the package first or even install it if needed.

```{r}
library(skimr)
plastic_waste %>% skim()
```

Summary statistics by continent.

```{r}
plastic_waste %>% 
  group_by(continent) %>% 
  skim()
```

# Exercises 3.

Let's start by taking a look at the distribution of plastic waste per capita in 2010.

```{r plastic_waste_per_cap-hist, eval=TRUE}
plastic_waste %>% 
  ggplot() +
  aes(x = plastic_waste_per_cap) +
  geom_histogram()
```

One country stands out as an unusual observation at the top of the distribution.

One way of identifying this country is to filter the data for countries where plastic waste per capita is greater than 3.5 kg/person.

```{r plastic_waste_per_cap-max, eval=TRUE}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```

Did you expect this result? You might consider doing some research on Trinidad and Tobago to see why plastic waste per capita is so high there, or whether this is a data error.

Another way of visualizing numerical data is using density plots.

Built a density plot below.

```{r plastic_waste_per_cap-dens}
plastic_waste %>% 
  ggplot() +
  aes(x = plastic_waste_per_cap) +
  geom_density()
```

And compare distributions across continents by colouring density curves by continent.

```{r plastic_waste_per_cap-dens-color}
plastic_waste %>% 
  ggplot() +
  aes(x = plastic_waste_per_cap, color = continent) +
  geom_density()
```

Transform the variable of plastic waste per capita to observe more informative density plot.

```{r}
plastic_waste %>% 
  mutate(plastic_waste_per_1000_pple = 
           plastic_waste_per_cap * 1000) %>% 
  ggplot() +
  aes(x = plastic_waste_per_1000_pple, 
      color = continent) +
  geom_density() + 
  scale_x_log10() + 
  labs(x = "Amount of plastic waste per 1000 people, kg / day")
```

# Exercises 4.

Built a box plot of plastic sate by continent

```{r}
plastic_waste %>% 
  mutate(plastic_waste_per_100_pple = 
           plastic_waste_per_cap * 100 * 365 / 1000) %>% 
  ggplot() +
  aes(x = plastic_waste_per_100_pple,
      y = continent, 
      color = continent) +
  geom_boxplot() + 
  scale_x_log10() + 
  labs(x = "Amount of plastic waste per 100 people, MT / year", 
       y = NULL)
```

# Exercises 5.

Build a scatter plot of plastic waste per capital and GDP per capital.

```{r}
plastic_waste %>% 
  mutate(plastic_waste_per_1000_pple = 
           plastic_waste_per_cap * 100 * 365 / 1000) %>% 
  ggplot() +
  aes(x = plastic_waste_per_1000_pple,
      y = gdp_per_cap, 
      color = continent,
      size = total_pop) +
  geom_point() + 
  scale_x_log10() + 
  scale_y_log10() + 
  labs(x = "Amount of plastic waste per 100 people, MT / year")
```
