---
title: "AE03-03 Correlation"
author: "Eduard Bukin"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---

## Setups

```{r setup, message=FALSE}
library(tidyverse)
library(readr)       # install.packages("readr")
library(readxl)      # install.packages("readxl")
library(janitor)     # install.packages("janitor")
library(skimr)       # install.packages("skimr")
library(lubridate)   # install.packages("lubridate")

ggplot2::theme_set(ggplot2::theme_minimal())

knitr::opts_chunk$set(out.width = 60)
```

## Data import and cleaning

### Importing and cleaning data in the wide format

Practically the same procedure that we did in the previous exercise.

```{r}
p_wd <- 
  read_excel("data/commodity-prices.xlsx", sheet =  "data") %>% 
  clean_names() %>% 
  rename(wheat = soft_red_winter_wheat_no_2_f_o_b_us_gulf_usd_per_mt, 
        maize = yellow_maize_no_2_f_o_b_us_gulf_usd_per_mt, 
        date = day_month_year, 
        oil = crude_oil_brent_usd_per_barrel, 
        urea = urea_f_o_b_black_sea_usd_per_mt) %>%
  slice(-1) %>% 
  mutate(
    oil = as.numeric(oil),
    wheat = as.numeric(wheat),
    maize = as.numeric(maize),
    urea = as.numeric(urea),
    date = convert_to_date(date)
  ) 

glimpse(p_wd)
```

### Converting data to the long format

```{r}
p_lg <- 
  p_wd %>% 
  pivot_longer(cols = c(oil:urea), 
               names_to = "var", 
               values_to = "price") %>% 
  arrange(date, var)
glimpse(p_lg)
```

## Example: Computing prices index with some base

Index of a variable with a base is a mathematical transformation of a variable, where each value of a variable is divided by the base value and then multiplied by 100.

$$
{I}_{i} = \frac{x_i}{x_{base}} \times 100
$$

This is easy in Excel, but tricky without it!

Let us demonstrate the logic of calculations based on a simple example of **2 commodities (wheat and urea)** and **few month** in **year 2010**.

#### Step 1. Filter a sub-sample

As discussed above, we want to:

-   `filter data, where year is 2010, month less than 6, var is "wheat" or "maize"`

Which identically translate into R code as:

-   `filter(data, year(date) == 2010, month(date) < 6, var %in% c("wheat", "maize"))`\`
-   `filter data, where year is 2010, month less than 6, var is "wheat" or "maize"`

```{r}
p_lg_sb <- 
  filter(p_lg, 
         year(date) == 2010, 
         month(date) < 6, 
         var %in% c("wheat", "maize")) %>% 
  arrange(var, date)
p_lg_sb
```

#### Step 2. Create variable with the base for indexing

Here we want to:

-   `for each groups of commodities "var", mutate variable "base", which is equal to "price" when month is equal to 1 and year is equal to 2010`.

In language of R this is:

```{r}
p_lg_sb %>% 
  group_by(var) %>%           # 1. for each groups of commodities "var"
  mutate(                     # 2. mutate
    base = ifelse(            # 3. variable "base", which is equal to
      month(date) == 1 &      # 5. when month is equal to 1 and
        year(date) == 2010,   # 6. year is equal to 2010
      price,                  # 4. "price"
      NA                      # 7. missing value is in other cases
      )
    )
```

#### Step 3. Make sure that base is the same for all observations in each group `var`

```{r}
p_lg_sb %>% 
  group_by(var) %>% 
  mutate(base = ifelse(month(date) == 1 & year(date) == 2010, price, NA)) %>% 
  tidyr::fill(base, .direction = "updown")
```

#### Step 4. Calculate index

```{r}
p_lg_sb %>% 
  group_by(var) %>% 
  mutate(base = ifelse(month(date) == 1 & year(date) == 2010, price, NA)) %>% 
  tidyr::fill(base, .direction = "updown") %>% 
  mutate(index = price / base * 100)
```

#### Optional example: calculate index, where base is average of month 2 and 3

```{r}
p_lg_sb %>%
  group_by(var) %>%
  mutate(base = ifelse(month(date) %in% c(2,3) & 
                         year(date) == 2010, 
                       price, 
                       NA)) %>% 
  mutate(base_full = mean(base, na.rm = TRUE)) %>% 
  mutate(index = price / base_full * 100)
```

Note, we specify `mean(…, na.rm = TRUE)` because when we compute mean, there are some missing observations in the variable. Mean of a vector with missing observation will return `NA`.

```{r}
mean(c(1, 2, 3, 4, NA))
```

If we specify parameter to ignore `NA` in the data, we will get a result:

```{r}
mean(c(1, 2, 3, 4, NA), na.rm = TRUE)
```

## Exercise 1. Compute prices index with base mean prices in 2010

Following the previous examples, let us compute:

-   Step 1. for each **group** of `var`;
-   Step 2. **mutate** `base_part`, which contains `price` **if** `year` is 2010 and `NA` in **else** cases;
-   Step 3. **mutate** `base` with the `mean()` value of `base_part` price with parameter `na.rm = TRUE`;
-   Step 4. **mutate** price `index` against such `base`;
-   Step 5. **ungroup** data
-   Step 6. **select** variables `date`, `var`, `price` and `index`

```{r}
# p_index <- 
#   p_lg %>%
#   ________() %>%                               # step 1.
#   ______(                                      # step 2.
#     base_part = 
#       ifelse(year(_____) == 2010, price, ____)
#     ) %>% 
#   ______(                                      # step 3.
#     base = ______(_______, na.rm = TRUE)
#     ) %>%  
#   mutate(index = ____ / _____ * 100) %>%       # step 4.
#   ungroup() %>%                                # step 5.
#   select(date, var, price, index)              # step 6.
# 
# glimpse(p_index)
```

## Exercise 2. Plot time series of indexes for wheat and urea

Before plotting we need to `filter var when it is %in% "wehat" or "urea"`;

Remember from the previous exercises: `ggplot()` + `aes()` + `geom_path()`

-   Use `labs(x = "", y = "", title = "")` to give meaningful labels to the plot.

```{r}
# p_index %>% 
#   ______(var %in% c("wheat", ______)) %>% 
#   ______() + 
#   aes(x = _____, y = _______, colour = var) + 
#   geom_path() + 
#   labs()
```

Answer the following questions:

-   Can we conclude, based on the plot, that surging prices of `urea` cause the `wheat` prices to surge?

-   What could be a theoretical explanation for this?

-   What could be the theoretical mechanism of `urea` prices effect on `wheat`?

## Exercise 3. Build a correlation table between price indices of different commodities

First, we need to convert our data to wide format again:

```{r}
# p_index_wd <-
#   p_index %>% 
#   pivot_wider(names_from = var, values_from = c(price, index))
# glimpse(p_index_wd)
```

To make a correlation table, we use package `correlation` and a function with the same name. We use summary to convert correlation table with extensive results to a compact matrix

-   We `select` only those variables, where names `contains()` string `index`.

```{r}
library(correlation)

# p_index_wd %>% 
#   select(contains("index")) %>% 
#   correlation() %>% 
#   summary()

# p_index_wd %>% 
#   select(contains("price")) %>% 
#   correlation() %>% 
#   summary()
```

-   Does this correlation coefficients suggests about causation if we assume that theory does justifies causal relationship?

Run the same correlations but without `summary()`. What are the differences?

```{r}
# p_index_wd %>% 
#   ______(_______("index")) %>% 
#   _________()

# ______ %>% 
#   ______(______("price")) %>% 
#   ______() 
```

## Exercise 4. Compute a first difference of indices with lag 1

First difference is a change of value in the next period, compared to the previous one. To compute it, we use function `lag()` and perform similar mutate operations.

#### Simple example of a first difference

Before, we computed index in the following way:

```{r}
p_lg_sb %>% 
  group_by(var) %>% 
  mutate(base = ifelse(month(date) == 1 & year(date) == 2010, price, NA)) %>% 
  tidyr::fill(base, .direction = "updown") %>% 
  mutate(index = price / base * 100)
```

let us `mutate()` the `index_fd` variable:

```{r}
p_lg_sb %>% 
  group_by(var) %>% 
  mutate(base = ifelse(month(date) == 1 & year(date) == 2010, price, NA)) %>% 
  tidyr::fill(base, .direction = "updown") %>% 
  mutate(index = price / base * 100) %>% 
  mutate(index_fd = index - lag(index))
```

Using simple example, let us compute the first difference of the index for the entire data.

```{r}
# p_index_fd <- 
#   ______ %>% 
#   group_by(______) %>% 
#   mutate(index_fd = ______ - lag(______)) %>% 
#   ungroup() %>% 
#   select(date, var, index_fd) %>% 
#   pivot_wider(names_from = var, 
#               values_from = c(index_fd))
#
# p_index_fd %>% 
#   glimpse()
```

## Exercise 5. Build a correlation table between first differences of indices for different commodities

As the same exercise before, we use `correlation` package and the same function.

```{r}
# p_index_fd %>% 
#   correlation() %>% 
#   summary()
```

-   Based on this results, does urea prices causes surges in the wheat prices?

-   What kind of causal relationship could be there?

## Exercise 6. Compute first differences with lag 2 and 3

```{r}
# p_index_fd_lags <- 
#   p_index %>% 
#   group_by(var) %>% 
#   mutate(fd = index - lag(index, 1)) %>% 
#   ungroup() %>% 
#   select(date, var, contains("fd")) %>% 
#   pivot_wider(names_from = var, 
#               values_from = c(contains("fd"))) %>% 
#   mutate(urea_fd1 = urea,
#          urea_fd2 = lag(urea, 2),
#          urea_fd3 = lag(urea, 3),
#          urea_fd4 = lag(urea, 4),
#          urea_fd5 = lag(urea, 5)
#   )
# 
# correlation(p_index_fd_lags) %>% summary()
```

## Solutions

```{r}
p_wd <- 
  read_excel("data/commodity-prices.xlsx", sheet =  "data") %>% 
  clean_names() %>% 
  rename(wheat = soft_red_winter_wheat_no_2_f_o_b_us_gulf_usd_per_mt, 
        maize = yellow_maize_no_2_f_o_b_us_gulf_usd_per_mt, 
        date = day_month_year, 
        oil = crude_oil_brent_usd_per_barrel, 
        urea = urea_f_o_b_black_sea_usd_per_mt) %>%
  slice(-1) %>% 
  mutate(
    oil = as.numeric(oil),
    wheat = as.numeric(wheat),
    maize = as.numeric(maize),
    urea = as.numeric(urea),
    date = convert_to_date(date)
  ) 

p_lg <- 
  p_wd %>% 
  pivot_longer(cols = c(oil:urea), 
               names_to = "var", 
               values_to = "price") %>% 
  arrange(var, date)
```

### Ex. 1

```{r}
p_index <- 
  p_lg %>%
  group_by() %>%                               # step 1.
  mutate(                                      # step 2.
    base_part = 
      ifelse(year(date) == 2010, price, NA)
    ) %>% 
  mutate(                                      # step 3.
    base = mean(base_part, na.rm = TRUE)
    ) %>%  
  mutate(index = price / base * 100) %>%       # step 4.
  ungroup() %>%                                # step 5.
  select(date, var, price, index)              # step 6.

glimpse(p_index)
```

### Ex. 2

```{r}
p_index %>% 
  filter(var %in% c("wheat", "urea")) %>%
  ggplot() + 
  aes(x = date, y = index, colour = var) + 
  geom_path() + 
  labs(x = "Date", y = "Price index, 2010 = 100",
       title = "Price indices of key commodities",
       colour = NULL)
```

### Ex. 3

```{r}
p_index_wd <-
  p_index %>% 
  pivot_wider(names_from = var, values_from = c(price, index))

library(correlation)

p_index_wd %>% 
  select(contains("index")) %>% 
  correlation() %>% 
  summary()

p_index_wd %>% 
  select(contains("price")) %>% 
  correlation() %>% 
  summary()
```

### Ex. 4

```{r}
p_index_fd <- 
  p_index %>% 
  group_by(var) %>% 
  mutate(index_fd = index - lag(index)) %>% 
  ungroup() %>% 
  select(date, var, index_fd) %>% 
  pivot_wider(names_from = var,
              values_from = c(index_fd))
```
