---
title: "Data import"
author: "Eduard Bukin"
date: "`r Sys.Date()`"
output: html_document
bibliography: ../../references.bib
---

## Setups

We will us `tidyverse` for data wrangling and `readr` and `readxl` for data import.

```{r setup, message=FALSE}
library(tidyverse)
library(readr)       # install.packages("readr")
library(readxl)      # install.packages("readxl")
```

# Data import from `.csv`

`.csv` is coma separated file.

Here is the example [NHIS 2009](https://raw.githubusercontent.com/EBukin/datasets4examples/main/datasets-raw/NHIS2009.csv) data used in the [@Angrist2014].

To load data we use `readr` package from tidyverse [readr](https://readr.tidyverse.org/).

We may use a visual import tool for data available in R studio. But remember to make the R code used to import data part of your analysis.

The challenge is to specify the right path to the file that we want to import


## Exercise 1. Simple CSV file

In the folder `data` there is a file `chicken.csv`. Load it into the object chick.

```{r}
library(readr)
# chick <- read_csv(____)
chick <- read_csv("data/chicken.csv")
```

Now inspect the data:

-   use `glimpse(____)`;
-   simply print the data set;

```{r}
read_csv()
```

Try function `summary(_____)`.

```{r}

```

What is the difference between summary and previous two ways of data exploration.
